---
title: "Comparison with cross-platform correlations from other studies"
author: "Noora Sissala"
date: "2025-04-09"
output: html_document
---


## Included studies


Comparison with MS-Olink correlations from:

- "Multiplatform Approach for Plasma Proteomics: Complementarity of Olink Proximity Extension Assay Technology to Mass Spectrometry-Based Protein Profiling" (Petrera et al., 2021) (DDA MS, DIA MS, and Olink data)
- - "Multi-platform proteomic analysis of Alzheimer’s disease cerebrospinal fluid and plasma reveals network biomarkers associated with proteostasis and the matrisome" (Dammer et al., 2022) (MS, Olink, and SomaScan data)


Comparison with MS-SomaScan correlations from:

- Dammer et al.


Comparison with Olink-SomaScan correlations from:

- Dammer et al.
- "Proteomic profiling platforms head to head: Leveraging genetics and clinical traits to compare aptamer- and antibody-based methods" (Katz et al, 2022),
- "Large-scale plasma proteomics comparisons through genetics and disease associations" (Eldjarn et al, 2023), 
-  "Synergistic insights into human health from aptamer- and antibody-based proteomic profiling" (Pietzner et al, 2021),
- "Comparison of Proteomic Measurements Across Platforms in the Atherosclerosis Risk in Communities (ARIC) Study" (Rooney et al., 2023),
- "Comparison of Proteomic Assessment Methods in Multiple Cohort Studies" (Raffield et al., 2020),
- "Dual-platform affinity proteomics identifies links between the recurrence of ovarian carcinoma and proteins released into the tumor microenvironment" (Finkernagel et al., 2019),
- "Stability and reproducibility of proteomic profiles in epidemiological studies: comparing the Olink and SOMAscan platforms" (Haslam et al., 2022).


Analysis of confidence tiers from:
- Eldjarn et al.,


## Load data


```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```


```{r, message = FALSE}

library(tidyverse)
library(readxl)
library(ggrepel)
library(ggside)
library(ggpp)
library(ggpubr)
library(patchwork)

source('code/utility_functions.R')
source('code/plotting_functions.R')
source('code/correlation_analysis_functions.R')

path <- path <- "results/correlation_comparisons"

if (!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
}

```

```{r, message=FALSE}

# Load correlations from this study
my_cors <- read.csv('results/correlation_analysis/MS-Olink_protein_cor_table.csv') |> 
  dplyr::rename(Correlation = Correlation_Spearman)

# Petrera et al.
petrera_olink_ann <- read_xlsx('data/external_data/pr0c00641_si_003.xlsx', 
                               range = 'A5:ABY8', col_names = F,
                               .name_repair = 'universal_quiet') |> 
  t() |> 
  as.data.frame()
colnames(petrera_olink_ann) <- petrera_olink_ann[1,]
petrera_olink_ann <- petrera_olink_ann[-1,]
rownames(petrera_olink_ann) <- NULL

petrera_LOD <-read_xlsx('data/external_data/pr0c00641_si_003.xlsx', 
                        range = 'A184:ABY186', col_names = F,
                        .name_repair = 'universal_quiet') |> 
  t() |> 
  as.data.frame()
colnames(petrera_LOD) <- make.names(petrera_LOD[1,])
petrera_LOD <- petrera_LOD[-1,]
rownames(petrera_LOD) <- NULL
petrera_LOD <- petrera_LOD |> 
  mutate(across(1:2, as.numeric))

petrera_NPX <- read_xlsx('data/external_data/pr0c00641_si_003.xlsx', 
                           range = 'A10:ABY182', col_names = F,
                         .name_repair = 'universal_quiet') |> 
  t() |> 
  as.data.frame()
colnames(petrera_NPX) <- petrera_NPX[1,]
petrera_NPX <- petrera_NPX[-1,]
rownames(petrera_NPX) <- NULL

petrera_olink <- bind_cols(petrera_olink_ann, petrera_LOD, petrera_NPX) |>  
  # OlinkIDs of duplicate assays that were removed in the paper
  filter(!OlinkID %in% c('OID00947', 'OID00666', 'OID01009', 'OID00684'))

petrera_dda_samples <- read_xlsx('data/external_data/pr0c00641_si_004.xlsx',
                                 sheet = 2, range = 'H1:FY2', col_names = F,
                                 .name_repair = 'universal_quiet') |> 
  t() |> 
  as.data.frame()
colnames(petrera_dda_samples) <- petrera_dda_samples[1,]
petrera_dda_samples <- petrera_dda_samples[-1,]
rownames(petrera_dda_samples) <- NULL

petrera_dda <- read_xlsx('data/external_data/pr0c00641_si_004.xlsx',
                         sheet = 2, skip = 2, .name_repair = 'universal_quiet') |> 
  dplyr::rename_with(\(x) str_extract(x, '(?<=\\.Normalized\\.\\.\\.).*(?=\\.\\.Sample)'),
                     contains('Sample', ignore.case = F)) |> 
  mutate(across(starts_with('F'), log2))  # log2-transform values
colnames(petrera_dda) <- ifelse(
  colnames(petrera_dda) %in% petrera_dda_samples$`ID (PD2.3)`,
  petrera_dda_samples$`Sample ID`[match(colnames(petrera_dda), petrera_dda_samples$`ID (PD2.3)`)],
  colnames(petrera_dda))


# Dammer et al.

dammer_soma_olink_cors <- read_xlsx(
  'data/external_data/13195_2022_1113_MOESM2_ESM.xlsx',
  sheet = '13.Olink.SomaScan.Corr.Plasma', skip = 3,
  .name_repair = 'universal_quiet') |> 
  dplyr::select(-1) |> 
  dplyr::rename(Olink = OlinkAssay,
                SomaScan = SomaAssay) |>
  separate_wider_delim(cols = c(Olink, SomaScan),
                       names = c('Gene.Name', 'UniProt'),
                       names_sep = '.',
                       delim = '|') |> 
  separate_wider_delim(cols = SomaScan.UniProt,
                       names = c('SomaScan.UniProt', 'SomaID'),
                       delim = '^',
                       too_few = 'align_start')
  
dammer_ms_olink_cors <- read_xlsx(
  'data/external_data/13195_2022_1113_MOESM2_ESM.xlsx',
  sheet = '15.Olink.MS.Corr.Plasma', skip = 3,
  .name_repair = 'universal_quiet') |> 
  dplyr::select(-1) |> 
  dplyr::rename(Olink = OlinkAssay,
                MS = Msprotein) |>
  separate_wider_delim(cols = c(Olink, MS),
                       names = c('Gene.Name', 'UniProt'),
                       names_sep = '.',
                       delim = '|')

dammer_ms_soma_cors <- read_xlsx(
  'data/external_data/13195_2022_1113_MOESM2_ESM.xlsx',
  sheet = '17.MS.SomaScan.Corr.Plasma', skip = 3,
  .name_repair = 'universal_quiet') |> 
  dplyr::select(-1) |> 
  dplyr::rename(MS = MSprotein,
                SomaScan = SomaAssay) |>
  separate_wider_delim(cols = c(MS, SomaScan),
                       names = c('Gene.Name', 'UniProt'),
                       names_sep = '.',
                       delim = '|') |>
  separate_wider_delim(cols = SomaScan.UniProt,
                       names = c('SomaScan.UniProt', 'SomaID'),
                       delim = '^',
                       too_few = 'align_start')

# Nilsson et al.
nilsson_cors <- read_xlsx(
  'data/external_data/004_corr_shared.xlsx',
  .name_repair = 'universal_quiet') |> 
  mutate(UniProt = str_remove(UnitedIDs, '.+_'),
         .after = 1)

# Katz et al
katz_cors <- read_xlsx('data/external_data/sciadv.abm5164_tables_s1_to_s6.xlsx',
                       sheet = 'Table S2', skip = 1,
                       .name_repair = 'universal_quiet') %>% 
  filter(!is.na(Soma.Target) & !is.na(Olink.Target)) |> 
  dplyr::rename(Cognate_Soma = Cognate...18, 
                Cis_Soma = Cis...19,
                Cognate_Olink = Cognate...26,
                Cis_Olink = Cis...27)

# Eldjarn et al
eldjarn_cors <- read_xlsx('data/external_data/41586_2023_6563_MOESM3_ESM.xlsx',
                          sheet = 'ST6_assay_correlations', skip = 3, 
                          .name_repair = 'universal_quiet', col_types = 'text') |> 
  mutate(across(7:17, as.numeric)) |> 
  filter(!is.na(olink_smpnorm_corr))

eldjarn_tiers <- read_xlsx('data/external_data/41586_2023_6563_MOESM3_ESM.xlsx',
                          sheet = 'ST29_protein_classification', skip = 2, 
                          .name_repair = 'universal_quiet', col_types = 'text') |> 
  filter(Targeted.Olink == 'Y', Targeted.Soma == 'Y') |> 
  mutate(olink.smp.corr = as.numeric(olink.smp.corr))
 

# Pietzner et al
pietzner_cors <- read_xlsx('data/external_data/41467_2021_27164_MOESM5_ESM.xlsx',
                           skip = 2, .name_repair = 'universal_quiet') |> 
  filter(!is.na(SomaScan.SeqID) & !is.na(OlinkID)) |> 
  dplyr::rename(Correlation.norm = Spearman.correlation.coefficient..normalised.,
                Correlation.nonnorm = Spearman.correlation.coefficient..non.normalised.) |> 
  mutate(across(1:2, \(x) str_remove(x, "^\\'")))

# Rooney et al
rooney_cors <- read_xlsx(
  'data/external_data/02 Proteomics validation Supp 20220815 SF1 updated.xlsx',
  sheet = 'ST3 soma v olink', range = 'A2:AN419',
  .name_repair = 'universal_quiet') |> 
  mutate(across(15:16, as.numeric)) |> # NAs introduced due to 3 values that are "NA"
  filter(!is.na(SOMAmer.SeqID) & !is.na(OlinkID)) |> 
  dplyr::rename(Correlation = Correlation.SomaScan.vs.Olink,
                Correlation.outliers.excl = Correlation.SomaScan.vs.Olink.outliers.excluded)

# Raffield et al.
raffield_cors <- read_xlsx(
  'data/external_data/pmic13292-sup-TableS9.xlsx')

# Finkeragel et al.
finkernagel_cors <- read_xlsx(
  'data/external_data/thnov09p6601s1.xlsx',
  sheet = 'Table S5', .name_repair = 'universal_quiet') |> 
  dplyr::rename(Gene.Name = ...1)

# Haslam et al.
haslam_cors <- read_xlsx(
  'data/external_data/pmic13529-sup-0001-tabless1-s17.xlsx',
  sheet = 'Supplemental Table S12', .name_repair = 'universal_quiet',
  skip = 2)

```



## Comparison with other MS-Olink studies

### Petrera et al.


#### Correlation calculation 

I matched proteins by UniProt ID and calculated Spearman correlations between DDA-MS and Olink data.

```{r, warning=FALSE}

source('code/correlation_analysis_functions.R')

petrera_dda_olink_cors <- cor.test.df(petrera_dda[-c(1,3:8)], petrera_olink[-c(1:2,4:7)],
                                      match_by = c('Accession', 'Uniprot ID'),
                                      cor_type = 'rowwise', cor_method = 'spearman')

nrow(petrera_dda_olink_cors)


```


#### Overall correlation distribution


All MS (DDA)-Olink correlations from the study:

```{r}

petrera_hist <- petrera_dda_olink_cors |> 
  mutate(Cor.interval = cor_intervals(Correlation)) |>
  cor_hist(table_y = 10, table_x = -0.4) +
  xlim(c(-0.4, 1)) +
  labs(title = 'Petrera et al.')

petrera_hist

```


#### Correlation of overlapping protein pairs

Matched proteins between studies by UniProt ID. 

```{r}

petrera_dda_overlap <- my_cors %>%
  inner_join(petrera_dda_olink_cors, by = join_by(UniProt == Accession),
             suffix = c('_Sissala', '_Petrera')) |> 
  mutate(Diff = Correlation_Sissala - Correlation_Petrera,
         Interval_Sissala = cor_intervals(Correlation_Sissala),
         Interval_Petrera = cor_intervals(Correlation_Petrera)) |>
  dplyr::rename(Sissala = Correlation_Sissala, Petrera = Correlation_Petrera)


nrow(petrera_dda_overlap)

```


```{r}

comp_cor_hist <- function(data, n_overlap, col1, col2) {
  # Function for correlation histogram
  
  data <- mutate(data, 
                 Study = paste(Study, 'et al.'),
                 Study.group = ifelse(Study == 'Sissala et al.', 'Current study', 'Previous study'),
                 Study.group = factor(Study.group, levels = c('Current study', 'Previous study')))
  
  median_df <- data |> 
    group_by(Study, Study.group) |>
    summarise(Median = median(Correlation, na.rm = T)) |> 
    ungroup() |> 
    add_row(Study = NA, Median = NA) |> 
    rowwise() |> 
    mutate(npcx = 'left',
           label = ifelse(is.na(Study), str_glue('N = {n_overlap}'),
                          str_glue('Median {Study}: {round(Median, 2)}'))) |>
    ungroup() |> 
    mutate(npcy = rev(calculate.label.npcy('top', 3)))
  
  # add p value to plot below the medians
  pval_position <- min(median_df$npcy) - 0.05
  pval <- wilcox.test(Correlation ~ Study, data = data)$p.value
  pval_label <- paste('Wilcoxon,', ifelse(pval < 0.001, 'p < 0.001', paste('p =', round(pval, 3))))
  
  data |> 
    ggplot() +
    geom_histogram(aes(Correlation, fill = Study.group, color = Study.group), 
                   binwidth = 0.05, position = 'identity',
                   alpha = 0.5, color = 'grey30', lwd = 0.3) +
    geom_text_npc(data = median_df[1:2,], aes(npcx = npcx, npcy = npcy, label = label, color = Study.group),
                  size = 6/.pt) +
    geom_text_npc(data = median_df[3,], aes(npcx = npcx, npcy = npcy, label = label),
                  size = 6/.pt, fontface = 'bold') +
    geom_vline(data = median_df[1:2,], aes(xintercept = Median, color = Study.group),
               lwd = 0.3, linetype = 'dashed') +
    theme_publ() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    labs(fill = 'Study', color = 'Study', x = 'Cross-platform correlation', y = 'Count') +
    geom_text_npc(
      data = data.frame(
        x = 'left', y = pval_position, label = pval_label
      ),
      aes(npcx = x, npcy = y, label = label), size = 6/.pt,
      inherit.aes = FALSE
    )
  
  
}


petrera_overlap_hist <- petrera_dda_overlap |> 
  pivot_longer(cols = c(Sissala, Petrera),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(petrera_dda_overlap))


petrera_overlap_hist


```


### Dammer et al.

As proteins were matched by the authors by gene name to calculate correlations, some correlations were between proteins with different UniProt IDs. I filtered the correlations to ones matched by UniProt ID. The authors calculated Pearson correlations, unlike the other studies, which used Spearman correlations.


#### Overall correlation distribution


Number and proportion of correlations matched by UniProt ID (not just gene name):

```{r}

# Percentage matched by UniProt from all correlations
dammer_ms_olink_cors |> 
  filter(MS.UniProt == Olink.UniProt) |> 
  nrow()

paste0(round(dammer_ms_olink_cors |> 
  filter(MS.UniProt == Olink.UniProt) |> 
  nrow() / nrow(dammer_ms_olink_cors) * 100, 2), '%')

```

Correlations for proteins matched by UniProt ID (N = 254):

```{r}

dammer_hist <- dammer_ms_olink_cors |> 
  filter(MS.UniProt == Olink.UniProt) |> 
  dplyr::rename(Correlation = all.cor) |> 
  mutate(Cor.interval = cor_intervals(Correlation)) |> 
  cor_hist(table_y = 47, table_x = -0.4) +
  xlim(c(-0.4, 1)) +
  labs(title = 'Dammer et al.')

dammer_hist

```

From here on analyses were performed on the proteins matched by UniProt ID.


#### Correlation of overlapping protein pairs

Matched proteins between studies by UniProt ID.

```{r}

dammer_ms_olink_overlap <- my_cors %>%
  inner_join(filter(dammer_ms_olink_cors, MS.UniProt == Olink.UniProt), 
             by = join_by(UniProt == MS.UniProt)) |> 
  mutate(Diff = Correlation - all.cor,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Dammer = cor_intervals(all.cor)) |> 
  dplyr::rename(Sissala = Correlation, Dammer = all.cor)

nrow(dammer_ms_olink_overlap)

```

```{r}


dammer_overlap_hist <- dammer_ms_olink_overlap |> 
  pivot_longer(cols = c(Sissala, Dammer),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(dammer_ms_olink_overlap))


dammer_overlap_hist


```


### Nilsson et al.


#### Overall correlation distribution


```{r}

nilsson_hist <- nilsson_cors |> 
  dplyr::rename(Correlation = Corr) |> 
  mutate(Cor.interval = cor_intervals(Correlation)) |> 
  cor_hist(table_y = 20) +
  labs(title = 'Nilsson et al.')

```

```{r}

nilsson_hist

```


#### Correlation of overlapping protein pairs

Matched proteins between studies by UniProt ID.

```{r}

nilsson_overlap <- my_cors %>%
  inner_join(nilsson_cors, by = 'UniProt',
             suffix = c('.Sissala', '.Nilsson')) |> 
  dplyr::rename(Sissala = Correlation, Nilsson = Corr) |>
  mutate(Diff = Sissala - Nilsson,
         Interval_Sissala = cor_intervals(Sissala),
         Interval_Nilsson = cor_intervals(Nilsson))

nrow(nilsson_overlap)

```


```{r}

nilsson_overlap_hist <- nilsson_overlap |> 
  pivot_longer(cols = c(Sissala, Nilsson),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(nilsson_overlap))

nilsson_overlap_hist


```



### Summary


```{r}

# Merge correlation data for overlapping protein pairs from the different studies
ms_olink_overlap_cors <- bind_rows(
  'Petrera et al.' = petrera_dda_overlap |> 
    dplyr::rename(Correlation_Sissala = Sissala,
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  N_Previous.study = N_Petrera,
                  Correlation_Previous.study = Petrera,
                  P.value_Previous.study = P.value,
                  Adj.p.value_Previous.study = Adj.p.value,
                  Cor.method_Previous.study = Cor.method,
                  Interval_Previous.study = Interval_Petrera) |>
    mutate(Cor.method_Sissala = 'Spearman'),
  'Dammer et al.' = dammer_ms_olink_overlap |> 
    dplyr::rename(N_Sissala = N,
                  Correlation_Sissala = Sissala,
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  N_Previous.study = all.nObs,
                  Correlation_Previous.study = Dammer,
                  P.value_Previous.study = all.p,
                  Interval_Previous.study = Interval_Dammer) |> 
    mutate(Cor.method_Sissala = 'Spearman',
           Cor.method_Previous.study = 'Pearson'),
  'Nilsson et al.' = nilsson_overlap |> 
    dplyr::rename(N_Sissala = N,
                  Correlation_Sissala = Sissala,
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Nilsson,
                  P.value_Previous.study = p,
                  Adj.p.value_Previous.study = p_adjust,
                  Interval_Previous.study = Interval_Nilsson) |>
    mutate(Cor.method_Sissala = 'Spearman',
           Cor.method_Previous.study = 'Spearman'),
  .id = 'Study'
) |> 
  dplyr::select(UniProt, Gene.Name, Description, 
                N_Sissala, Correlation_Sissala, P.value_Sissala, Adj.p.value_Sissala, 
                Interval_Sissala, Cor.method_Sissala, Study, N_Previous.study, 
                Correlation_Previous.study, P.value_Previous.study, 
                Adj.p.value_Previous.study, Cor.method_Previous.study, 
                Interval_Previous.study)



```


```{r}

write.csv(ms_olink_overlap_cors, file = file.path(path, 'MS-Olink_cor_comparison.csv'), row.names = FALSE)

```


```{r, fig.height=5}

# add titles
petrera_overlap_hist <- petrera_overlap_hist +
  labs(title = 'Petrera et al.', x = 'MS-Olink correlation')

dammer_overlap_hist <- dammer_overlap_hist +
  labs(title = 'Dammer et al.', x = 'MS-Olink correlation')

nilsson_overlap_hist <- nilsson_overlap_hist +
  labs(title = 'Nilsson et al.', x = 'MS-Olink correlation')


guide_area() + petrera_overlap_hist + dammer_overlap_hist + nilsson_overlap_hist +
  plot_spacer() +
  plot_layout(design = 'AA\nBC\nDE', heights = c(0.1, 1, 1),
              guides = 'collect')

```

```{r}

ggsave(filename = 'MS-Olink_cors_overlappingProteins.pdf',
       path = path,
       width = 18.3,
       height = 15,
       units = 'cm')

```



## Comparison with MS-SomaScan studies


### Dammer et al.

As proteins were matched by the authors by gene name to calculate correlations, some correlations were between proteins with different UniProt IDs. I filtered the correlations to ones matched by UniProt ID. 


#### Overall correlation distribution

Number and proportion of correlations matched by UniProt ID (not just gene name):

```{r}

# Percentage matched by UniProt from all correlations
dammer_ms_soma_cors |> 
  filter(MS.UniProt == SomaScan.UniProt) |> 
  nrow()

paste0(round(dammer_ms_soma_cors |> 
  filter(MS.UniProt == SomaScan.UniProt) |> 
  nrow() / nrow(dammer_ms_soma_cors) * 100, 2), '%')

```


Correlations for proteins matched by UniProt ID:

```{r}

dammer_ms_soma_cors |>
  filter(MS.UniProt == SomaScan.UniProt) |> 
  ggplot.histogram(all.cor, binwidth = 0.05, fill = 'steelblue',
                   alpha = 0.8, color = 'steelblue', 
                   median_line_params = list(lwd = 0.3),
                   median_text_params = list(size = 6/.pt),
                   xlab = 'Correlation',
                   title = 'Dammer et al. MS-SomaScan correlations') +
    theme_publ()

```

From here on analyses were performed on the proteins matched by UniProt ID.


#### Correlation of overlapping protein pairs

```{r}

dammer_ms_soma_overlap <- my_cors %>%
  inner_join(filter(dammer_ms_soma_cors, MS.UniProt == SomaScan.UniProt), 
             by = join_by(UniProt == MS.UniProt)) |> 
  dplyr::rename(Sissala = Correlation, Dammer = all.cor) |> 
  mutate(Diff = Sissala - Dammer,
         Interval_Sissala = cor_intervals(Sissala),
         Interval_Dammer = cor_intervals(Dammer))

nrow(dammer_ms_soma_overlap)

```


```{r}

dammer_ms_soma_overlap_hist <- dammer_ms_soma_overlap |> 
  pivot_longer(cols = c(Sissala, Dammer),
               names_to = 'Study', 
               values_to = 'Correlation') |>
  comp_cor_hist(n_overlap = nrow(dammer_ms_soma_overlap)) +
  labs(title = 'Dammer et al.', x = 'MS-SomaScan correlation') 

guide_area() / dammer_ms_soma_overlap_hist + plot_layout(guides = 'collect',
              heights = c(0.1,1))
```

```{r}

ggsave(filename = 'MS-SomaScan_cors_overlappingProteins.pdf',
       path = path,
       width = 8.7,
       height = 6.5,
       units = 'cm')

```


Save results:

```{r}

ms_soma_overlap_cors <- my_cors %>%
  inner_join(filter(dammer_ms_soma_cors, MS.UniProt == SomaScan.UniProt), 
             by = join_by(UniProt == MS.UniProt)) |> 
  mutate(Diff = Correlation - all.cor,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Previous.study = cor_intervals(all.cor),
         Study = 'Dammer et al.',
         Cor.method_Sissala = 'Spearman',
         Cor.method_Previous.study = 'Pearson') |>
  dplyr::rename(N_Sissala = N,
                Correlation_Sissala = Correlation,
                P.value_Sissala = P.value_Spearman,
                Adj.p.value_Sissala = Adj.p.value_Spearman,
                N_Previous.study = all.nObs,
                Correlation_Previous.study = all.cor,
                P.value_Previous.study = all.p) |> 
  dplyr::select(UniProt, Gene.Name, Description, N_Sissala, Correlation_Sissala,
                 P.value_Sissala, Adj.p.value_Sissala, Interval_Sissala, 
                 Cor.method_Sissala, Study, N_Previous.study, Correlation_Previous.study,
                 P.value_Previous.study, Cor.method_Previous.study, Interval_Previous.study)

```

```{r}

write.csv(ms_soma_overlap_cors, file.path(path, 'MS-SomaScan_cor_comparison.csv'),
          row.names = F)

```



## Comparison with Olink-SomaScan studies

### Katz et al


#### Overall correlation distribution

```{r}

ggplot.histogram(katz_cors, rho, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Katz et al. Olink-SomaScan correlations') +
  theme_publ()

```


#### Correlation of overlapping protein pairs

```{r}

katz_overlap <- my_cors %>%
  inner_join(katz_cors, by = join_by(UniProt == UniProt.x)) |> 
  mutate(Diff = Correlation - rho,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Katz = cor_intervals(rho)) |> 
  dplyr::rename(Sissala = Correlation, Katz = rho)

nrow(katz_overlap)

```


```{r}

katz_overlap_hist <- katz_overlap |> 
  pivot_longer(cols = c(Sissala, Katz),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(katz_overlap))

katz_overlap_hist

```


### Eldjarn et al


I chose to use the SMP-normalized correlations from Eldjarn et al. here, as presented in the main figures of the paper.

#### Overall correlation distribution

```{r}

ggplot.histogram(eldjarn_cors, olink_smpnorm_corr, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Eldjarn et al. Olink-SomaScan correlations') +
  theme_publ()


```


#### Correlation of overlapping protein pairs

```{r}

eldjarn_overlap <- my_cors %>%
  inner_join(eldjarn_cors, by = join_by(UniProt == uniprot)) |> 
  mutate(Diff = Correlation - olink_smpnorm_corr,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Eldjarn = cor_intervals(olink_smpnorm_corr)) |> 
  dplyr::rename(Sissala = Correlation, Eldjarn = olink_smpnorm_corr)

nrow(eldjarn_overlap)

```


```{r}

eldjarn_overlap_hist <- eldjarn_overlap |> 
  pivot_longer(cols = c(Sissala, Eldjarn),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(eldjarn_overlap))

eldjarn_overlap_hist

```


#### Cross-platform correlations by confidence tier


```{r}

# Merge correlations from the current study and Eldjarn et al.
tier_cors <- my_cors |> 
  inner_join(eldjarn_tiers, by = c('UniProt', 'OlinkID')) |>
  mutate(olink.smp.corr = as.numeric(olink.smp.corr)) |> 
  # For duplicated assays in Olink/SomaScan in Eldjarn,
  # keep the one with the highest correlation
  arrange(-olink.smp.corr) |> 
  distinct(UniProt, .keep_all = T) |> 
  dplyr::select(1:8, P.value_Spearman, Adj.p.value_Spearman, olink.smp.corr, 
                olink.smp.pval, cis.pQTL.olink, PAV.olink, eQTL.olink, 
                cis.pQTL.soma, PAV.soma, eQTL.soma, confidence.tier) |> 
  dplyr::rename(Correlation_Sissala = Correlation, 
                P.value_Sissala = P.value_Spearman,
                Adj.p.value_Sissala = Adj.p.value_Spearman,
                Correlation_Eldjarn = olink.smp.corr,
                P.value_Eldjarn = olink.smp.pval,
                cis.pQTL.olink_Eldjarn = cis.pQTL.olink,
                PAV.olink_Eldjarn = PAV.olink,
                eQTL.olink_Eldjarn = eQTL.olink,
                cis.pQTL.soma_Eldjarn = cis.pQTL.soma,
                PAV.soma_Eldjarn = PAV.soma,
                eQTL.soma_Eldjarn = eQTL.soma,
                confidence.tier_Eldjarn = confidence.tier) |> 
  # Which platform each cis-pQTL was identified on
  mutate(cis.pQTL.platform_Eldjarn = case_when(
    cis.pQTL.olink_Eldjarn == 'Y' & cis.pQTL.soma_Eldjarn == 'Y' ~ 'Both',
    cis.pQTL.olink_Eldjarn == 'Y' & cis.pQTL.soma_Eldjarn == 'N' ~ 'Olink only',
    cis.pQTL.soma_Eldjarn == 'Y' & cis.pQTL.olink_Eldjarn == 'N' ~ 'SomaScan only',
    cis.pQTL.soma_Eldjarn == 'N' & cis.pQTL.olink_Eldjarn == 'N' ~ 'None'),
    cis.pQTL.platform_Eldjarn = factor(
      cis.pQTL.platform_Eldjarn, 
      levels = c('Both', 'Olink only', 'SomaScan only', 'None')),
    # Add column for high correlation in both studies
    High.corr.in.both = Correlation_Sissala > 0.7 & Correlation_Eldjarn > 0.7)
                
# Color palette for confidence tiers
tier_palette <- desaturate(darken(brewer.pal(3, 'Set2')[c(1,3,2)]), -0.2)


```


```{r}

# Calculate median cross-platform correlation by confidence tier
median_df <- tier_cors |> 
  group_by(confidence.tier_Eldjarn) |> 
  summarise(Median.Eldjarn = median(Correlation_Eldjarn, na.rm = TRUE),
            Median.Sissala = median(Correlation_Sissala, na.rm = T))

tier_cors |>
  ggplot.scatterplot(Correlation_Eldjarn, Correlation_Sissala, 
                     Color_var = confidence.tier_Eldjarn,
                   xlab = 'Olink-SomaScan correlation (Eldjarn et al.)',
                   ylab = 'MS-Olink correlation (Sissala et al.)', 
                   title = 'MS-Olink vs. Olink-SomaScan correlations',
                   alpha = 0.8) +
  # Add density plots showing the distribution of cross-platform correlations to the sides
  geom_xsidedensity(aes(fill = confidence.tier_Eldjarn), alpha = 0.5) +
  geom_ysidedensity(aes(fill = confidence.tier_Eldjarn), alpha = 0.5) +
  geom_xsidevline(data = median_df,
                  mapping = aes(xintercept = Median.Eldjarn, color = confidence.tier_Eldjarn),
                  linetype = 'dashed', show.legend = F) +
  geom_ysidehline(data = median_df, 
                  mapping = aes(yintercept = Median.Sissala, color = confidence.tier_Eldjarn),
                  linetype = 'dashed', show.legend = F) +
  theme_publ()  +
  theme(ggside.axis.text = element_blank(),
        ggside.axis.line = element_blank(),
        ggside.axis.ticks = element_blank()) +
  labs(fill = 'Confidence tier (Eldjarn et al.)', color = 'Confidence tier (Eldjarn et al.)',
       x = 'Olink-SomaScan correlation (Eldjarn et al.)',
       y = 'MS-Olink correlation (Sissala et al.)') +
  geom_abline(linetype = 'dashed', lwd = 0.4)  +
  scale_color_manual(values = tier_palette,
                     aesthetics = c('color', 'fill'))

median_df

```

```{r}

ggsave(filename = 'Eldjarn_confidence_tier_cors.pdf',
       path = path,
       width = 8.7,
       height = 7.5,
       units = 'cm')

```

```{r}

write.csv(tier_cors, file.path(path, 'Eldjarn_confidence_tier_cors.csv'), row.names = F)

```


Boxplot of cross-platform correlation by tiers:

```{r}

p1 <- ggplot.jitterbox(tier_cors, confidence.tier_Eldjarn, Correlation_Sissala, show.legend = F) +
  geom_pwc(method = 't.test', tip.length = 0.02, label.size = 6/.pt,
           p.adjust.method = 'fdr', label = 'p.adj.signif', step.increase = 0.1, vjust = 0.5) +
  labs(x = 'Confidence tier (Eldjarn et al.)', y = 'MS-Olink correlation (Sissala et al.)') +
  scale_color_manual(values = tier_palette,
                     aesthetics = c('color', 'fill')) +
  labs(title = 'MS-Olink correlation')

p2 <- ggplot.jitterbox(tier_cors, confidence.tier_Eldjarn, Correlation_Eldjarn, show.legend = F) +
  geom_pwc(method = 't.test', tip.length = 0.02, label.size = 6/.pt,
           p.adjust.method = 'fdr', label = 'p.adj.signif', step.increase = 0.1, vjust = 0.5) +
  labs(x = 'Confidence tier (Eldjarn et al.)', y = 'Olink-SomaScan correlation (Eldjarn et al.)') +
  scale_color_manual(values = tier_palette,
                     aesthetics = c('color', 'fill')) +
  labs(title = 'Olink-SomaScan correlation')

guide_area() / (p1 | p2) + plot_layout(guides = 'collect', height = c(0.01,1))

```

```{r}

ggsave(filename = 'cor_by_tier_boxplot.pdf',
       path = path,
       width = 11.3,
       height = 7,
       units = 'cm')

```


Cross-platform correlation by pQTLs:

```{r}

tier_cors |> 
  ggplot.jitterbox(cis.pQTL.platform_Eldjarn, Correlation_Sissala, show.legend = F) +
  labs(x = 'cis pQTL in (Eldjarn et al.)', y = 'MS-Olink correlation (Sissala et al.)') +
  theme(ggside.axis.text = element_blank(),
        ggside.axis.line = element_blank(),
        ggside.axis.ticks = element_blank(),
        ggside.panel.scale = 0.2) +
  geom_pwc(method = 't.test', p.adjust.method = 'fdr', 
           tip.length = 0.02, label.size = 6/.pt, label = 'p.adj.signif', step.increase = 0.1) +
  stat_group_counts(label.x = 'cis.pQTL.platform')
  
  

```

```{r}

ggsave(filename = 'MS-Olink_cor_by_Eldjarn_pQTLs.pdf',
       path = path,
       width = 8.7,
       height = 6.5,
       units = 'cm')


```


```{r}

tier_cors |> pivot_longer(
  cols = c(Correlation_Sissala, Correlation_Eldjarn), 
  names_to = c('.value', 'Study'),
  names_sep = '_') |>
  mutate(Study = ifelse(
    Study == 'Sissala', 'Sissala et al.', 'Eldjarn et al.'),
    Study = factor(Study, levels = c('Sissala et al.', 'Eldjarn et al.'))) |> 
  ggplot.jitterbox(confidence.tier_Eldjarn, Correlation, Color_var = Study) +
  geom_pwc(method = 'wilcox_test', label.size = 6/.pt, tip.length = 0.02,
           label = 'p.adj.signif', p.adjust.method = 'fdr') +
  theme_publ() +
  labs(x = 'Confidence tier (Eldjarn et al.)', y = 'Cross-platform correlation', fill = 'Study')

```

```{r}

ggsave(filename = 'cor_by_tier_Sissala_vs_Eldjarn.pdf',
       path = path,
       width = 8.7,
       height = 6.5,
       units = 'cm')

```



### Pietzner et al


The authors calculated correlations based on SMP-normalized and non-normalized data. I chose to use the normalized correlations here.


#### Overall correlation distribution

```{r}

ggplot.histogram(pietzner_cors, Correlation.norm, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Pietzner et al. Olink-SomaScan correlations') +
  theme_publ()


```


#### Correlation of overlapping protein pairs

Proteins were matched between studies by UniProt IDs.

```{r}

pietzner_overlap <- my_cors %>%
  inner_join(pietzner_cors, by = join_by(UniProt == UniprotID)) |> 
  mutate(Diff = Correlation - Correlation.norm,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Pietzner = cor_intervals(Correlation.norm)) |> 
  dplyr::rename(Sissala = Correlation, Pietzner = Correlation.norm)

nrow(pietzner_overlap)

```


```{r}

 
pietzner_overlap_hist <- pietzner_overlap |> 
  pivot_longer(cols = c(Sissala, Pietzner),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(pietzner_overlap))

pietzner_overlap_hist

```



### Rooney et al.


#### Overall correlation distribution

```{r}

ggplot.histogram(rooney_cors, Correlation, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Rooney et al. Olink-SomaScan correlations') +
  theme_publ()


```


#### Correlation of overlapping protein pairs


Matching was done using UniProt IDs.

```{r}

rooney_overlap <- my_cors %>%
  inner_join(rooney_cors, by = join_by(UniProt == UniprotID),
             suffix = c('_Sissala', '_Rooney')) |> 
  mutate(Diff = Correlation_Sissala - Correlation_Rooney,
         Interval_Sissala = cor_intervals(Correlation_Sissala),
         Interval_Rooney = cor_intervals(Correlation_Rooney)) |> 
  dplyr::rename(Sissala = Correlation_Sissala, Rooney = Correlation_Rooney)

nrow(rooney_overlap)

```

```{r}

 
rooney_overlap_hist <- rooney_overlap |> 
  pivot_longer(cols = c(Sissala, Rooney),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(rooney_overlap))

rooney_overlap_hist


```


### Raffield et al.


#### Overall correlation distribution

```{r}

ggplot.histogram(raffield_cors, rs, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Raffield et al. Olink-SomaScan correlations') +
  theme_publ()

```

#### Correlation of overlapping protein pairs

```{r}

raffield_overlap <- my_cors %>%
  inner_join(raffield_cors, by = 'UniProt') |> 
  mutate(Diff = Correlation - rs,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Raffield = cor_intervals(rs)) |> 
  dplyr::rename(Sissala = Correlation, Raffield = rs)

nrow(raffield_overlap)

```

```{r}

 
raffield_overlap_hist <- raffield_overlap |> 
  pivot_longer(cols = c(Sissala, Raffield),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(raffield_overlap))

raffield_overlap_hist


```


### Finkernagel et al.

The authors calculated both Pearson and Spearman correlations. I chose to use the Spearman correlations here as the other studies used Spearman correlations.

#### Overall correlation distribution

```{r}

ggplot.histogram(finkernagel_cors, Spearman.rho, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Finkernagel et al. Olink-SomaScan correlations') +
  theme_publ()


```

#### Correlation of overlapping protein pairs

```{r}

finkernagel_overlap <- my_cors %>%
  inner_join(finkernagel_cors, by = 'Gene.Name') |> 
  mutate(Diff = Correlation - Spearman.rho,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Finkernagel = cor_intervals(Spearman.rho)) |> 
  dplyr::rename(Sissala = Correlation, Finkernagel = Spearman.rho)

nrow(finkernagel_overlap)

```


```{r}

finkernagel_overlap_hist <- finkernagel_overlap |> 
  pivot_longer(cols = c(Sissala, Finkernagel),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(finkernagel_overlap))

finkernagel_overlap_hist

```


### Haslam et al.



#### Overall correlation distribution

```{r}

ggplot.histogram(haslam_cors, r, binwidth = 0.05, fill = 'steelblue',
                 alpha = 0.8, color = 'steelblue', 
                 median_line_params = list(lwd = 0.3),
                 median_text_params = list(size = 6/.pt),
                 xlab = 'Correlation',
                 title = 'Haslam et al. Olink-SomaScan correlations') +
  theme_publ()

```

#### Correlation of overlapping protein pairs

```{r}

haslam_overlap <- my_cors %>%
  inner_join(haslam_cors, by = join_by(UniProt == UniprotID)) |> 
  mutate(Diff = Correlation - r,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Haslam = cor_intervals(r)) |> 
  dplyr::rename(Sissala = Correlation, Haslam = r)

nrow(haslam_overlap)

```

```{r}

haslam_overlap_hist <- haslam_overlap |> 
  pivot_longer(cols = c(Sissala, Haslam),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(haslam_overlap))

haslam_overlap_hist

```


### Dammer et al.


#### Overall correlation distribution


Proteins matched by UniProt ID:

```{r}

dammer_soma_olink_cors |>
  filter(SomaScan.UniProt == Olink.UniProt) |>
  nrow() / nrow(dammer_soma_olink_cors) * 100

```

```{r}

dammer_soma_olink_cors |>
  filter(SomaScan.UniProt == Olink.UniProt) |>
  ggplot.histogram(all.cor, binwidth = 0.05, fill = 'steelblue',
                   alpha = 0.8, color = 'steelblue', 
                   median_line_params = list(lwd = 0.3),
                   median_text_params = list(size = 6/.pt),
                   xlab = 'Correlation',
                   title = 'Dammer et al. Olink-SomaScan correlations') +
    theme_publ()

```


#### Correlation of overlapping protein pairs


```{r}

dammer_olink_soma_overlap <- my_cors %>%
  inner_join(dammer_soma_olink_cors, by = c('UniProt' = 'SomaScan.UniProt')) |> 
  mutate(Diff = Correlation - all.cor,
         Interval_Sissala = cor_intervals(Correlation),
         Interval_Dammer = cor_intervals(all.cor)) |> 
  dplyr::rename(Sissala = Correlation, Dammer = all.cor)

nrow(dammer_olink_soma_overlap)


```


```{r}

dammer_olink_soma_overlap_hist <- dammer_olink_soma_overlap |> 
  pivot_longer(cols = c(Sissala, Dammer),
               names_to = 'Study', 
               values_to = 'Correlation') |> 
  comp_cor_hist(n_overlap = nrow(dammer_olink_soma_overlap))

dammer_olink_soma_overlap_hist

```


### Summary

```{r}

# Merge all correlation data
olink_soma_overlap_cors <- bind_rows(
  'Katz et al.' = katz_overlap |> 
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Katz,
                  P.value_Previous.study = p...11),
  'Eldjarn et al.' = eldjarn_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Eldjarn,
                  P.value_Previous.study = olink_smpnorm_pval),
  'Pietzner et al.' = pietzner_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Pietzner),
  'Rooney et al.' = rooney_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Rooney,
                  P.value_Previous.study = P.value.Correlation.SomaScan.vs.Olink),
  'Raffield et al.' = raffield_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Raffield,
                  N_Previous.study = n),
  'Finkernagel et al.' = finkernagel_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Finkernagel),
  'Dammer et al.' = dammer_olink_soma_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Dammer,
                  P.value_Previous.study = all.p,
                  N_Previous.study = all.nObs),
  'Haslam et al.' = haslam_overlap |>
    dplyr::rename(N_Sissala = N, 
                  Correlation_Sissala = Sissala, 
                  P.value_Sissala = P.value_Spearman,
                  Adj.p.value_Sissala = Adj.p.value_Spearman,
                  Correlation_Previous.study = Haslam),
  .id = 'Study') |> 
  mutate(Interval_Sissala = cor_intervals(Correlation_Sissala),
         Cor.method_Sissala = 'Spearman',
         Cor.method_Previous.study = case_when(
           Study == 'Dammer et al.' ~ 'Pearson',
           TRUE ~ 'Spearman'),
         Interval_Previous.study = cor_intervals(Correlation_Previous.study)) |> 
  dplyr::select(UniProt, Gene.Name, Description, N_Sissala, Correlation_Sissala, 
                P.value_Sissala, Adj.p.value_Sissala, Interval_Sissala,
                Cor.method_Sissala, Study, N_Previous.study, Correlation_Previous.study, 
                P.value_Previous.study, Interval_Previous.study, Cor.method_Previous.study)

```


```{r}

write.csv(olink_soma_overlap_cors, file.path(path, 'Olink-SomaScan_cor_comparison.csv'), 
          row.names = F)

```


```{r, fig.height=8, fig.width=5}

# Add titles
dammer_olink_soma_overlap_hist <- dammer_olink_soma_overlap_hist +
  labs(title = 'Dammer et al.') 

eldjarn_overlap_hist <- eldjarn_overlap_hist +
  labs(title = 'Eldjarn et al.')

finkernagel_overlap_hist <- finkernagel_overlap_hist +
  labs(title = 'Finkernagel et al.')

haslam_overlap_hist <- haslam_overlap_hist +
  labs(title = 'Haslam et al.')

katz_overlap_hist <- katz_overlap_hist +
  labs(title = 'Katz et al.')

pietzner_overlap_hist <- pietzner_overlap_hist +
  labs(title = 'Pietzner et al.')

raffield_overlap_hist <- raffield_overlap_hist +
  labs(title = 'Raffield et al.')

rooney_overlap_hist <- rooney_overlap_hist +
  labs(title = 'Rooney et al.')


guide_area()  +
dammer_olink_soma_overlap_hist + eldjarn_overlap_hist + finkernagel_overlap_hist + 
  haslam_overlap_hist + katz_overlap_hist + pietzner_overlap_hist + raffield_overlap_hist +
  rooney_overlap_hist +
  plot_layout(design = 'AA\nBC\nDE\nFG\nHI',
              heights = c(0.1,1,1,1,1),
              guides = 'collect')

```

```{r}

ggsave(filename = 'Olink-SomaScan_cors_overlappingProteins.pdf',
       path = path,
       width = 18.3, 
       height = 21,
       units = 'cm')

```



