 ---
title: "CV analysis"
author: "Noora Sissala"
date: "2025-04-07"
output: html_document
---

## Read in data

```{r setup}

library(tidyverse)
library(scales)
library(ggpubr)
library(patchwork)
library(ggpp)
library(rstatix)

```

```{r}

# Result directory
path <- 'results/CV_analysis'

if (!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
}

source('code/utility_functions.R')
source('code/plotting_functions.R')
source('code/CV_analysis_functions.R')

```


```{r}

# Read in data
ms_data_withReplicates <- read_MS_data(withReplicateSamples = TRUE)
ms_data <- read_MS_data(withReplicateSamples = FALSE)

olink_data_withControls <- read_OlinkExplore_data(withControlSamples = TRUE,
                                                  withBelowLOD = TRUE, # Keep <LOD values
                                                  withQCWarning = TRUE,
                                                  withReplicateAssays = TRUE)

olink_data_withControls_wide <- olink_long_to_wide(
  olink_data_withControls, 
  id_cols = c('UniProt', 'OlinkID'))

olink_data <- read_OlinkExplore_data(withControlSamples = FALSE,
                                     withBelowLOD = TRUE, # Keep <LOD values
                                     withQCWarning = TRUE,
                                     withReplicateAssays = TRUE)

olink_data_wide <- olink_long_to_wide(
  olink_data, 
  id_cols = c('UniProt', 'OlinkID'))

metadata_ms <- read_metadata(withReplicates = TRUE, samples = 'MS')
metadata_overlap <- read_metadata(withReplicates = FALSE, samples = 'overlapping')

protein_metadata <- read.csv('data/metadata/protein_metadata.csv') |> 
  filter(UniProt %in% ms_data$UniProt | OlinkID %in% unique(olink_data$OlinkID))


# Load HPA data
hpa_data <- read.delim('data/processed_data/HPA_v24_clean.txt')

# Get blood concentration data
conc_data <- hpa_data %>% 
  filter(!is.na(Blood.conc.ngmL)) %>% 
  distinct(Uniprot, Blood.conc.ngmL, .keep_all = TRUE) |> 
  dplyr::select(Uniprot, Blood.conc.ngmL)


```


## Technical variation

Technical variation in MS and Olink Explore measurements was assessed per protein by calculating coefficients of variation (CV) on duplicate measurements.

### Number of duplicate measurements per protein

```{r, message=FALSE}

ms_duplicates <- metadata_overlap %>% 
  filter(Duplicate) %>% 
  pull(Sample.ID)

n_ms <- lapply(ms_duplicates, function(id) {
  pair_cols <- grep(id, colnames(ms_data_withReplicates), value = TRUE)
  no_na_in_pair <- rowSums(is.na(ms_data_withReplicates[pair_cols])) == 0
  data.frame(no_na_in_pair)
})
n_ms <- bind_cols(n_ms, .name_repair = 'universal_quiet')
n_ms <- rowSums(n_ms)

ggplot() +
  geom_bar(aes(x = n_ms)) +
  labs(x = 'Number of duplicates per protein (MS)', y = 'Count') +
  scale_x_continuous(breaks = seq(0, 6, 1))


```

```{r}

n_olink <- olink_data_withControls_wide %>% 
  dplyr::select(contains('Control')) %>% 
  apply(1, function(x) sum(!is.na(x)))
n_olink <- ifelse(n_olink == 2, 1, 0)

ggplot() +
  geom_bar(aes(x = n_olink)) +
  labs(x = 'Number of duplicates per protein (Olink)', y = 'Count')


```


### Technical CVs

#### All proteins

```{r}

# Calculate coefficients of variation (CV) on replicate samples
technical_CVs <- calc_tech_CV(ms_data = ms_data_withReplicates, 
                              olink_data = olink_data_withControls)

technical_CVs_MS <- technical_CVs |>
  filter(Platform == 'MS') |>
    dplyr::select(-OlinkID) |> 
  left_join(protein_metadata, by = 'UniProt') |> 
  dplyr::select(Platform, UniProt, Gene.Name, Description, Assay, OlinkID, 
                Panel, Technical.CV, Analyzed)

technical_CVs_Olink <- technical_CVs |>
  filter(Platform == 'Olink') |>
  left_join(protein_metadata, by = c('UniProt', 'OlinkID')) |>
  dplyr::select(Platform, UniProt, Gene.Name, Description, Assay, OlinkID, 
                Panel, Technical.CV, Analyzed)

technical_CVs <- bind_rows(technical_CVs_MS, technical_CVs_Olink)

technical_CVs |> 
  group_by(Platform) |>
  summarize(N_analyzed = sum(Analyzed),
            N_analyzed_unique = n_distinct(UniProt[Analyzed]),
            N_total = n(),
            N_total_unique = n_distinct(UniProt),
            Perc_analyzed = N_analyzed/N_total*100,
            Perc_analyzed_unique = N_analyzed_unique/N_total_unique*100)

```

```{r}

write.csv(technical_CVs, file = paste0(path, '/technical_CVs.csv'), row.names = FALSE)

```


```{r}

technical_CVs <- technical_CVs |>
  filter(!is.na(Technical.CV))

```


```{r}

CV_summary <- technical_CVs %>% 
  group_summary(summary_var = Technical.CV, Platform)

CV_summary

```

Histogram of technical CVs:

```{r}


stat_df <- CV_summary |>  
  mutate(y = c(49,47),
         npcy = c(0.95,0.9))

pval <- wilcox.test(Technical.CV ~ Platform, data = technical_CVs)$p.value
pval_label <- format_pval_label(pval)

ggplot(technical_CVs, aes(x = Technical.CV, fill = Platform, color = Platform)) +
  geom_histogram(aes(y = after_stat(
    c(count[group==1]/sum(count[group==1]),
      count[group==2]/sum(count[group==2]))*100)), 
    position = 'identity', alpha = 0.5, binwidth = 5, boundary = 0) +
  scale_fill_manual(values = TDP_palette[c(1,4)],
                    aesthetics = c('color', 'fill')) +
  labs(x = 'Technical CV (%)', y = 'Proportion of proteins (%)',
       title = 'Technical CVs') +
  theme_publ() +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0,100,10)) +
    # add point and line for median and IQR
  geom_point(
    data = stat_df,
    mapping = aes(x = Median, y = y, color = Platform),
    size = 1.5) +
  geom_segment(
    data = stat_df,
    mapping = aes(x = Q1, xend = Q3, color = Platform, y = y),
    linewidth = 0.5) +
  # add p.value
  geom_text_npc(data = data.frame(
    npcx = 'right', npcy = 0.95, label = pval_label), 
    mapping = aes(npcx = npcx, npcy = npcy, label = label),
    inherit.aes = FALSE, size = 6/.pt, hjust = 1, parse = TRUE) +
  theme(legend.position = 'inside',
        legend.position.inside = c(0.85, 0.6))

```

```{r}

ggsave(filename = 'technical_CV_hist.pdf',
       path = path,
       width = 8.3,
       height = 6,
       units = 'cm')

```

Bar plot of technical CV intervals:

```{r}

intervals <- c('0-5', '>5-10', '>10-15', '>15-20', '>20-50', '>50-100', '>100')

N_total <- technical_CVs |> 
  filter(!is.na(Technical.CV)) |> 
  count(Platform) |> 
  dplyr::rename(N.total = n)

CV_per_interval <- technical_CVs |> 
  filter(!is.na(Technical.CV)) |>
  group_by(Platform) |> 
  mutate(Interval = cut(Technical.CV, breaks = c(0, 5, 10, 15, 20, 50, 100, Inf),
                        labels = intervals, right = TRUE, include.lowest = TRUE)) |> 
  count(Platform, Interval) |> 
  left_join(N_total) |>
  mutate(Interval = factor(Interval, levels = intervals),
         Perc = n/N.total*100,
         Cumulative = cumsum(Perc))

CV_per_interval |> 
  ggplot(aes(x = Interval, y = Perc, fill = Platform)) +
  geom_col(position = 'dodge') +
  geom_line(aes(y = Cumulative, color = Platform, group = Platform,
                linetype = '')) +
    geom_point(aes(y = Cumulative), color = 'grey30', size = 1,
             stroke = 0.3,
             shape = 21, show.legend = F) +
  scale_fill_manual(values = TDP_palette[c(1,4)], aesthetics = c('fill', 'color')) +
  labs(x = 'Technical CV (%)', y = 'Proportion of proteins (%)', 
       linetype = 'Cumulative % of proteins') +
  theme_publ()  +
  scale_y_continuous(expand = expansion(c(0,0.05))) 

```

```{r}

ggsave(filename = 'technical_CV_intervals_barplot.pdf',
       path = path,
       width = 8.3,
       height = 6,
       units = 'cm')

```

```{r}

CV_per_interval

```



#### Overlapping proteins

```{r}

technical_CVs_overlap <- technical_CVs |> 
  filter(UniProt %in% intersect(ms_data$UniProt, olink_data$UniProt)) |> 
  pivot_wider(id_cols = UniProt,
              names_from = Platform, 
              values_from = Technical.CV) |> 
  filter(!is.na(MS) & !is.na(Olink)) |> 
  pivot_longer(cols = c(MS, Olink),
                names_to = 'Platform',
                values_to = 'Technical.CV')


stat_df2 <- group_summary(technical_CVs_overlap, Technical.CV, Platform) |>  
  mutate(y = c(59,57),
         npcy = c(0.95,0.9))

pval2 <- wilcox.test(Technical.CV ~ Platform, data = technical_CVs_overlap)$p.value
pval_label2 <- format_pval_label(pval2)

ggplot(technical_CVs_overlap, aes(x = Technical.CV, fill = Platform, color = Platform)) +
  geom_histogram(aes(y = after_stat(
    c(count[group==1]/sum(count[group==1]),
      count[group==2]/sum(count[group==2]))*100)), 
    position = 'identity', alpha = 0.5, binwidth = 5, boundary = 0) +
  scale_fill_manual(values = TDP_palette[c(1,4)],
                    aesthetics = c('color', 'fill')) +
  labs(x = 'Technical CV (%)', y = 'Proportion of proteins (%)') +
  theme_publ() +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0,100,10)) +
    # add point and line for median and IQR
  geom_point(
    data = stat_df2,
    mapping = aes(x = Median, y = y, color = Platform),
    size = 1.5) +
  geom_segment(
    data = stat_df2,
    mapping = aes(x = Q1, xend = Q3, color = Platform, y = y),
    linewidth = 0.5) +
  # add p.value
  geom_text_npc(data = data.frame(
    npcx = 'right', npcy = 0.95, label = pval_label2), 
    mapping = aes(npcx = npcx, npcy = npcy, label = label),
    inherit.aes = FALSE, size = 6/.pt, hjust = 1, parse = TRUE)

pval2


```

```{r}

ggsave(filename = 'technical_CV_hist_overlappingProteins.pdf',
       path = path,
       width = 8.3,
       height = 6,
       units = 'cm')

```



## Technical CVs vs. concentration in blood


Calculate concentration intervals:

```{r}

conc_intervals <- conc_data |> 
  filter(Uniprot %in% protein_metadata$UniProt) |>
  pull(Blood.conc.ngmL) |>
  quantile(probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = T)

interval_labels <- c('1st quantile', '2nd quantile', '3rd quantile', '4th quantile')

CV_per_conc <- technical_CVs |> 
  filter(!is.na(Technical.CV)) |> 
  left_join(conc_data, by = join_by(UniProt == Uniprot)) |>
  filter(!is.na(Blood.conc.ngmL)) |> 
  mutate(Conc.interval = cut(
    Blood.conc.ngmL, conc_intervals, include.lowest = T, labels = interval_labels),
    Conc.interval = factor(Conc.interval, levels = interval_labels))

```


Plot boxplots:

```{r}


stat_test <- CV_per_conc |> 
  group_by(Platform) |>
  wilcox_test(Technical.CV ~ Conc.interval, p.adjust.method = 'fdr',
              comparisons = list(
                c('1st quantile', '2nd quantile'), c('2nd quantile', '3rd quantile'),
                c('3rd quantile', '4th quantile'))) |> 
  add_xy_position(y.trans = log10)

set.seed(1)
CV_per_conc |> 
  ggplot.jitterbox(Conc.interval, Technical.CV) +
  facet_wrap(~Platform, scales = 'free_x', axes = 'all') +
  scale_y_log10(labels = log10_format()) +
  scale_fill_manual(values = brewer.pal(9, 'YlGnBu')[c(8:4)],
                    aesthetics = c('fill', 'color')) +
  small_logticks('l') +
  labs(x = 'Concentration in blood (ng/mL)', y = 'Technical CV (%)') +
  theme_publ() +
  stat_pvalue_manual(stat_test, size = 6/.pt, tip.length = 0.01, inherit.aes = F,
                     step.increase = 0.02, bracket.nudge.y = 0.1, 
                     step.group.by = 'Platform', label = 'p.adj.signif') +
  theme(legend.position = 'none')


```

```{r}

ggsave(filename = 'technicalCV_bloodConc_boxplot.pdf',
       path = path,
       width = 18.3,
       height = 7.5,
       units = 'cm')

```


## Technical CVs vs missing values


Calculate the proportion of missing values per protein:

```{r}

intervals <- c('0', '(0,25]', '(25,50]', '(50,75]', '(75,100]')

# Below LOD in Olink data
below_LOD <- olink_data %>%
  group_by(UniProt) %>%
  summarize(N.missing = sum(NPX < LOD),
            Perc.missing = N.missing/n()*100,
            N = n(),
            Interval = as.character(cut(
              Perc.missing, breaks = seq(-25,100,25), labels = intervals, 
              include.lowest = F, right = T)),
            Interval = factor(Interval, levels = intervals))

# Missing values in MS data
ms_NAs <- ms_data_withReplicates %>%
  pivot_longer(cols = -UniProt,
               values_to = 'Value',
               names_to = 'Sample.ID') %>%
  group_by(UniProt, .drop = FALSE) %>%
  summarize(N.missing = sum(is.na(Value)),
            Perc.missing = sum(is.na(Value))/n()*100,
            N = n(),
            Interval = as.character(cut(
              Perc.missing, breaks = seq(-25,100,25), labels = intervals, 
              include.lowest = F, right = T)),
            Interval = factor(Interval, levels = intervals))

# Merge data
CV_per_belowLOD <- technical_CVs |> 
  filter(Platform == 'Olink', !is.na(Technical.CV)) |>
  left_join(below_LOD, by = 'UniProt') 

CV_per_NAs <- technical_CVs |> 
  filter(Platform == 'MS', !is.na(Technical.CV)) |>
  left_join(ms_NAs, by = 'UniProt')

CV_per_missing <- bind_rows(CV_per_belowLOD, CV_per_NAs)

```


Plot boxplots:

```{r}

# Test for differences in CVs between intervals
stat_test1 <- CV_per_missing |> 
  filter(Platform == 'MS') |> 
  group_by(Platform) |>
  wilcox_test(Technical.CV ~ Interval, p.adjust.method = 'fdr',
              comparisons = list(
                c('0', '(0,25]'), c('(0,25]', '(25,50]'), c('(25,50]', '(50,75]'))) |> 
  add_xy_position(y.trans = log10)

stat_test2 <- CV_per_missing |>
  filter(Platform == 'Olink') |>
  group_by(Platform) |>
  wilcox_test(Technical.CV ~ Interval, p.adjust.method = 'fdr',
              comparisons = list(
                c('0', '(0,25]'), c('(0,25]', '(25,50]'), c('(25,50]', '(50,75]'), c('(50,75]', '(75,100]'))) |> 
  add_xy_position(y.trans = log10)

stat_test <- bind_rows('MS' = stat_test1, 'Olink' = stat_test2, .id = 'Platform')

# Plot boxplots
set.seed(1)
CV_per_missing |> 
  ggplot.jitterbox(Interval, Technical.CV, Color_var = Interval) +
  facet_wrap(~Platform, scales = 'free_x', axes = 'all') +
  scale_y_log10(labels = log10_format()) +
  scale_fill_manual(values = brewer.pal(9, 'YlGnBu')[c(8:4)],
                    aesthetics = c('fill', 'color')) +
  small_logticks('l') +
  labs(x = 'Proportion of missing values (%)', y = 'Technical CV (%)') +
  theme_publ() +
  stat_pvalue_manual(stat_test, size = 6/.pt, tip.length = 0.01, inherit.aes = F,
                     step.increase = 0.02, bracket.nudge.y = 0.1, step.group.by = 'Platform') +
  theme(legend.position = 'none')

```

```{r}

ggsave(filename = 'technicalCV_vs_missingValues.pdf',
       path = path,
       width = 18.3,
       height = 7.5,
       units = 'cm')

```

## Technical CVs at peptide level

```{r}

peptides <- read.delim('data/processed_data/peptide_table_all.txt')
replicates <- c('S034' , 'S049' , 'S058' , 'S071' , 'S089' , 'S098')

protein_metadata_MS <- protein_metadata |>
  filter(!is.na(Gene.Name))

peptide_metadata <- read.csv('data/metadata/peptide_metadata_MS.csv', 
                             row.names = 1) |> 
  filter(UniProt.ID %in% protein_metadata_MS$UniProt) # peptides mapping uniquely to the detected proteins

peptide_CVs <- peptides %>% 
  filter(Peptide.label %in% peptide_metadata$Peptide.label) |> # only peptides mapping to the detected proteins
  dplyr::select(Peptide.label, contains(replicates)) %>% 
  pivot_longer(cols = -Peptide.label,
               names_to = 'Sample.ID',
               values_to = 'Value') %>% 
  mutate(Sample.ID = str_remove(Sample.ID, 'a|b')) %>% 
  group_by(Peptide.label, Sample.ID) %>% 
  summarize(Technical.CV = CV.log2(Value)) %>% 
  group_by(Peptide.label) %>% 
  summarize(Technical.CV = mean(Technical.CV, na.rm = T)) |> 
  mutate(Analyzed = !is.na(Technical.CV) & !is.nan(Technical.CV),
         Technical.CV = ifelse(Technical.CV > 100, 100, Technical.CV),
         Technical.CV = ifelse(is.nan(Technical.CV), 100, Technical.CV)) |> 
# add metadata
left_join(peptide_metadata, by = 'Peptide.label') |> 
dplyr::select(Peptide.label, Peptide.sequence, UniProt.ID, Gene.Name, Description, 
              Technical.CV, Analyzed)


```

```{r}


stat_df <- peptide_CVs |>  
  group_summary(summary_var = Technical.CV) |> 
  mutate(y = 35)

ggplot(peptide_CVs, aes(x = Technical.CV)) +
  geom_histogram(aes(y = after_stat(count/sum(count))*100), 
    position = 'identity', binwidth = 5, boundary = 0, fill = TDP_palette[1], 
    color = TDP_palette[1], alpha = 0.5) +
  labs(x = 'Technical CV (%)', y = 'Proportion of peptides (%)',
       title = 'MS technical CVs at peptide level') +
  theme_publ() +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0,100,10)) +
    # add point and line for median and IQR
  geom_point(
    data = stat_df,
    mapping = aes(x = Median, y = y),
    size = 1.5, color = TDP_palette[1]) +
  geom_segment(
    data = stat_df,
    mapping = aes(x = Q1, xend = Q3, y = y),
    linewidth = 0.5, color = TDP_palette[1])

```

```{r}

ggsave(filename = 'peptide_CV_hist.pdf',
       path = path,
       width = 8.3,
       height = 6,
       units = 'cm')

```

## Session info

```{r}

writeLines(capture.output(sessionInfo()), file.path(path, 'session_info.txt'))

```

