---
title: "Differential abundance analysis - females vs. males"
author: "Noora Sissala"
date: "2025-04-07"
output: html_document
---

## Read in data

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```

```{r}

library(tidyverse)
library(RColorBrewer)
library(patchwork)
library(ggvenn)
library(eulerr)
library(ggpubr)
library(ggside)
library(rstatix)
library(scales)

source('code/utility_functions.R')
source('code/plotting_functions.R')
source('code/DAA_functions.R')

# Result directory
path <- 'results/DAA'

if (!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
}

#### Read in data ####

# Olink data with LOD values set to NA
olink_data_clean <- read_OlinkExplore_data(withControlSamples = FALSE,
                                           withReplicateAssays = FALSE,
                                           withBelowLOD = FALSE, # set <LOD values to NA
                                           withQCWarning = TRUE, # keep proteins/samples with QC warning
                                           removeAllLOD = TRUE) # remove proteins with all values <LOD

olink_data_clean_wide <- olink_long_to_wide(olink_data_clean, id_cols = 'UniProt')

# Olink data with no LOD values
olink_noNA <- na.omit(olink_data_clean_wide)


# MS data
ms_data <- read_MS_data(withIS = FALSE, # remove internal standards
                        withReplicateSamples = FALSE) # remove replicate samples


# Calculate proportion of NAs per protein on full MS data set
ms_data_all <- read_MS_data(withIS = FALSE, 
                            withReplicateSamples = TRUE) 

# Remove proteins with more than 50% missing values
ms_max50NA <- na_omit_prop(ms_data_all, 0.5, ignore_cols = 'UniProt')
# Remove proteins with any missing values
ms_noNA <- na.omit(ms_data)

# Overlapping proteins
overlap_noNA <- intersect(olink_noNA$UniProt, ms_noNA$UniProt)
length(overlap_noNA)

# Metadata
metadata <- read_metadata(withReplicates = FALSE, samples = 'overlapping')

# Protein annotations
protein_metadata <- read.csv('data/metadata/protein_metadata.csv') %>% 
  filter(OlinkID %in% unique(olink_data_clean$OlinkID) | UniProt %in% ms_data$UniProt)

# Color palette
daa_pal <- setNames(c(TDP_palette[c(1,4, 13)], 'grey50'),
                    c('MS only', 'Olink only', 'Both', 'None'))


```

## Differential abundance analysis (DAA)

Differential abundance analysis was performed between males and females as an example. Analysis was performed on overlapping proteins with no missing values in MS or Olink data (N = 569). 


```{r}

# Define groups
male <- metadata[metadata$Sex == 'Male', 'Sample.ID']
female <- metadata[metadata$Sex == 'Female', 'Sample.ID']

```

Function for the analysis:

```{r}

DAA <- function(ms_data, olink_data, test_grp, ref_grp, path, filename,
                         protein_metadata, p_adjust = TRUE) {
  
  library(tidyverse)
  library(ggplot2)
  library(patchwork)
  library(ggpp)
  
  
  # Perform DAA for MS data
  ms_res <- ms_data %>% 
    rowwise.t.test(test_grp = test_grp, ref_grp = ref_grp) %>% 
    mutate(variable = ms_data$UniProt) %>% 
    dplyr::rename(UniProt = variable)
  
  if (!p_adjust) {
    ms_res <- ms_res |> 
      mutate(adj.p = p.value)
  }
  
  # Perform DAA for Olink data
  olink_res <- olink_data %>% 
    rowwise.t.test(test_grp = test_grp, ref_grp = ref_grp) %>% 
    mutate(variable = olink_data$UniProt) %>% 
    dplyr::rename(UniProt = variable) |> 
    rowwise()
  
  if (!p_adjust) {
    olink_res <- olink_res |> 
      mutate(adj.p = p.value)
  }
  
  # Combine results
  daa_res <- bind_rows('MS' = ms_res, 'Olink' = olink_res, .id = 'Platform') |> 
    left_join(protein_metadata, by = 'UniProt') |> 
    dplyr::select(Platform, UniProt, Gene.Name, Description, Assay, OlinkID, Panel,
                  everything()) |> 
    mutate(Significance = ifelse(adj.p < 0.05, 'S', 'NS'))
  
  # Save result table
  write.csv(daa_res, paste0(path, '/DAA_sex_res_', filename, '.csv'))
  
  # Wide format
    daa_res_wide <- daa_res %>% 
    pivot_wider(id_cols = UniProt,
                names_from = Platform,
                names_sep = '_',
                values_from = c(statistic, log2FC, p.value, adj.p)) %>% 
    filter(!is.na(log2FC_MS) & !is.na(log2FC_Olink)) |>
    # Create column indicating which platform the protein is significant in
    mutate(Significance = case_when(adj.p_MS < 0.05 & adj.p_Olink < 0.05 ~ 'Both',
                           adj.p_MS < 0.05 & adj.p_Olink >= 0.05 ~ 'MS only',
                           adj.p_MS >= 0.05 & adj.p_Olink < 0.05 ~ 'Olink only',
                           TRUE ~ 'None'),
           Alpha = ifelse(Significance == 'None', 'None', 'Other'),
           Significance = factor(Significance, levels = c('Both', 'MS only', 'Olink only', 'None'))) 
  
  # Add labels to be used in volcano plots (if MS gene name is lacking, use Olink assay name)
    ms_res <- ms_res |> 
    left_join(protein_metadata, by = 'UniProt') |> 
    mutate(Gene.Name = ifelse(is.na(Gene.Name), Assay, Gene.Name))
  
  olink_res <- olink_res |>
    left_join(protein_metadata, by = 'UniProt') |>
    mutate(Gene.Name = ifelse(is.na(Gene.Name), Assay, Gene.Name))

  
  ## Volcano plots
  
  # Select proteins to label
  ms_labels <- full_join(
    slice_max(filter(ms_res, adj.p < 0.05), abs(log2FC), n = 10),
    slice_min(filter(ms_res, adj.p < 0.05), p.value, n = 10),
    by = colnames(ms_res))
  
  # Plot volcano
  ms_volcano <- DAA_volcano(ms_res, ms_labels) + labs(title = 'MS')
  
  # Select proteins to label
  olink_labels <- full_join(
    slice_max(filter(olink_res, adj.p < 0.05), abs(log2FC), n = 10),
    slice_min(filter(olink_res, adj.p < 0.05), p.value, n = 10),
    by = colnames(olink_res))
  
  # Plot volcano
  olink_volcano <- DAA_volcano(olink_res, olink_labels) + labs(title = 'Olink')
  
  # Combine plots
  volcano_plot <- ms_volcano + olink_volcano + 
    plot_layout(guides = 'collect', axes = 'collect') &
    theme(legend.position = 'top')
  
  ggsave(plot = volcano_plot,
         filename = paste0('DAA_sex_volcano_', filename, '.pdf'),
         path = path,
         width = 12,
         height = 7,
         units = 'cm')
  
  print(volcano_plot)
  
  # Venn diagram of DAPs
  ms_sig <- filter(ms_res, adj.p < 0.05) %>% pull(UniProt)
  olink_sig <- filter(olink_res, adj.p < 0.05) %>% pull(UniProt)
  
  # Calculate set sizes
  n_ms <- length(setdiff(ms_sig, olink_sig))
  n_olink <- length(setdiff(olink_sig, ms_sig))
  n_overlap <- length(intersect(ms_sig, olink_sig))
  
  # Plot and save Venn diagram
  pdf(file = paste0(path, '/DAA_sex_venn_', filename, '.pdf'),
      width = 5*0.393701, height = 5*0.393701)
  
    venn_diag <- plot(euler(
      combinations = c('MS' = n_ms,
                       'Olink' = n_olink,
                       'MS&Olink' = n_overlap)),
      fills = list(fill = TDP_palette[c(1,4)], alpha = 0.5), 
      edges = list(lwd = 2.5, col = TDP_palette[c(1,4)]),
      quantities = list(fontsize = 7, font = 'regular', fontfamily = 'Helvetica'), 
      labels = list(fontsize = 8, font = 'regular', fontfamily = 'Helvetica'))
  
    # Adjust plot to smaller size
    venn_diag$vp$width <- unit(0.8, "npc")
    venn_diag$vp$height <- unit(0.8, "npc")
    
    print(venn_diag)
    
  dev.off()
    
  print(venn_diag)
  
  ## Agreement of log2FC values
  
  # Percent agreement of log2FC values for all tested proteins
  perc_agreement <- daa_res |> 
    pivot_wider(id_cols = UniProt,
                names_from = Platform,
                values_from = log2FC) |> 
    filter(!is.na(MS) & !is.na(Olink)) |>
    mutate(Agreement = sign(MS) == sign(Olink)) |> 
    summarize(Perc.agreement = mean(Agreement, na.rm = T) * 100)
  
  # Scatter plot of log2FC agreement (all proteins)
  log2fc_plot1 <- ggplot(
    daa_res_wide, aes(x = log2FC_MS, y = log2FC_Olink, color = Significance, alpha = Alpha)) +
    geom_point(size = 0.75) +
    geom_quadrant_lines(lwd = 0.3) +
    geom_abline(lwd = 0.3) +
    geom_text_npc(
      data = perc_agreement, 
      aes(npcx = 0.65, npcy = 0.25, 
          label = paste0(
            "Categorical\nagreement: ", round(Perc.agreement), '%')),
      hjust = 0, vjust = 1, size = 6/.pt) +
    scale_alpha_manual(values = c(0.4, 0.9), guide = 'none') +
    scale_color_manual(values = daa_pal) +
    labs(x = 'log2-fold change (MS)', 
         y = 'log2-fold change (Olink)',
         color = 'Significant in',
         title = 'All overlapping proteins') +
    theme_publ() +
    theme(legend.position = 'top',
          plot.margin = margin(t = 5.5, b = 5.5, l = 5.5, r = 6.5, 'pt'))
  
  # Percent agreement of log2FC values for significant proteins only
  sig_any <- daa_res_wide |> 
    filter(Significance != 'None') |> 
    pull(UniProt)
  
  perc_agreement2 <- daa_res |> 
    filter(UniProt %in% sig_any) |>
    pivot_wider(id_cols = UniProt,
                names_from = Platform,
                values_from = log2FC) |> 
    filter(!is.na(MS) & !is.na(Olink)) |>
    mutate(Agreement = sign(MS) == sign(Olink)) |> 
    summarize(Perc.agreement = mean(Agreement, na.rm = T) * 100)
  
  # Scatter plot of log2FC agreement (significant proteins)
  log2fc_plot2 <- daa_res_wide |> 
    filter(UniProt %in% sig_any) |>
    ggplot.scatterplot(log2FC_MS, log2FC_Olink, Color_var = Significance, size = 0.75, alpha = 0.9) +
    scale_color_manual(values = daa_pal) +
    geom_quadrant_lines(lwd = 0.3) +
    geom_abline(lwd = 0.3) +
    geom_text_npc(
      data = perc_agreement2, 
      aes(npcx = 0.65, npcy = 0.25, 
          label = paste0(
            "Categorical\nagreement: ", round(Perc.agreement), '%')),
      hjust = 0, vjust = 1, size = 6/.pt) +
    labs(x = 'log2-fold change (MS)', 
         y = 'log2-fold change (Olink)',
         color = 'Significant in',
         title = 'Differentially abundant proteins') +
    theme_publ() +
    theme(legend.position = 'none',
          plot.margin = margin(t = 5.5, b = 5.5, l = 6.5, r = 5.5, 'pt'))
    
  # Combine plots
  log2fc_plot <- (log2fc_plot1 + log2fc_plot2) + 
    plot_layout(axis_titles = 'collect')
  log2fc_plot <- guide_area() / log2fc_plot + 
    plot_layout(heights = c(0.1,1), guides = 'collect') &
    guides(color = guide_legend(override.aes = list(size = 1.5)))
  
  ggsave(log2fc_plot,
         filename = paste0('DAA_sex_log2FC_agreement_', filename, '.pdf'),
         path = path,
         width = 12,
         height = 6.5,
         units = 'cm')
  
  print(log2fc_plot)
  
  return(list('DAA_res' = daa_res,
              'DAA_res_wide' = daa_res_wide))
  
}


```


### Overlapping proteins - no missing values

```{r}

daa_res_noNA_overlap <- DAA(
  filter(ms_noNA, UniProt %in% overlap_noNA),
  filter(olink_noNA, UniProt %in% overlap_noNA),
  test_grp = female,
  ref_grp = male,
  path = path,
  filename = 'noNA_overlappingProteins',
  protein_metadata = protein_metadata)

```


### Technical CVs by significance in DAA

```{r}

technical_CVs <- read.csv('results/CV_analysis/technical_CVs.csv')

hpa_data <- read.delim('data/processed_data/HPA_v24_clean.txt')

conc_data <- hpa_data %>% 
  filter(!is.na(Blood.conc.ngmL)) %>% 
  distinct(Uniprot, Blood.conc.IM.pgL, Blood.conc.MS.pgL, .keep_all = TRUE) |> 
  dplyr::select(Uniprot, Blood.conc.MS.pgL, Blood.conc.IM.pgL, Blood.conc.ngmL)

daa_res_noNA_overlap_wide <- daa_res_noNA_overlap$DAA_res_wide |> 
  left_join(technical_CVs |> 
              pivot_wider(id_cols = c(UniProt, Gene.Name),
                          names_from = Platform,
                          values_from = Technical.CV) |> 
              dplyr::rename(Technical.CV_MS = MS, Technical.CV_Olink = Olink),
            by = 'UniProt') |>
  left_join(conc_data, by = join_by(UniProt == Uniprot))

```

Technical CV by significance:

```{r}

plot_data <- daa_res_noNA_overlap_wide |> 
  filter(Significance != 'None', !is.na(Technical.CV_MS), !is.na(Technical.CV_Olink))

p1 <- ggplot.jitterbox(plot_data, Significance, Technical.CV_MS) +
  stat_compare_means(comparisons = list(c('MS only', 'Olink only'), c('Both', 'Olink only')),
                     label.y = c(30, 33), size = 6/.pt, tip.length = 0.01,
                     label = 'p.signif') +
  ylim(c(0,35)) +
  labs(x = 'Significant in', y = 'Technical CV (MS)') 
  

p2 <- ggplot.jitterbox(plot_data, Significance, Technical.CV_Olink) +
  ylim(c(0,35)) +
  stat_compare_means(comparisons = list(c('MS only', 'Olink only'), c('Both', 'MS only')),
                     label.y = c(30, 33), size = 6/.pt, tip.length = 0.01,
                     label = 'p.signif') +
  labs(x = 'Significant in', y = 'Technical CV (Olink)') 

p1 + p2 + plot_layout(guides = 'collect') &
  theme_publ() &
  theme(legend.position = 'top') &
  scale_color_manual(values = daa_pal, aesthetics = c('fill', 'color'))
  

```

```{r}

ggsave(filename = 'technical_CV_by_significance.pdf',
       path = path,
       width = 8.7,
       height = 6.5,
       units = 'cm')

```

### Fold change by significance in DAA

```{r}

plot_data <- daa_res_noNA_overlap_wide |> 
  filter(Significance != 'None') |> 
  mutate(absFC_MS = abs(log2FC_MS),
         absFC_Olink = abs(log2FC_Olink))

p1 <- ggplot.jitterbox(plot_data, Significance, absFC_MS) +
  stat_compare_means(comparisons = list(c('MS only', 'Olink only'), c('Both', 'Olink only')),
                     size = 6/.pt, tip.length = 0.01, label = 'p.signif') +
  labs(x = 'Significant in', y = 'log2-fold change (MS)') 
  

p2 <- ggplot.jitterbox(plot_data, Significance, absFC_Olink) +
  stat_compare_means(comparisons = list(c('MS only', 'Olink only'), c('Both', 'MS only')),
                     size = 6/.pt, tip.length = 0.01, label = 'p.signif') +
  labs(x = 'Significant in', y = 'log2-fold change (Olink)') 

p1 + p2 + plot_layout(guides = 'collect') &
  theme_publ() &
  theme(legend.position = 'top') &
  scale_color_manual(values = daa_pal, aesthetics = c('fill', 'color'))
  

```

```{r}

ggsave(filename = 'log2FC_vs_significance.pdf',
       path = path,
       width = 8.7,
       height = 6.5,
       units = 'cm')

```


### Fold change comparison


```{r}

daa_res_noNA_overlap$DAA_res_wide %>% 
  pivot_longer(cols = c(log2FC_MS, log2FC_Olink),
               names_to =  c('.value', 'Platform'),
               names_sep = '_') %>% 
  ggplot.density(log2FC, Fill_var = Platform, median_text = F, median_line = F) +
  scale_fill_manual(values = TDP_palette[c(1,4)], aesthetics = c('color', 'fill')) +
  theme_publ() +
  labs(title = '', x = 'log2-fold change')


ggsave(filename = 'log2FC_density_noNA_overlappingProteins.pdf',
       path = path,
       width = 8.7,
       height = 6,
       units = 'cm')

```

```{r}

daa_res_noNA_overlap$DAA_res_wide %>% 
  filter(Significance != 'None') %>% 
  pivot_longer(cols = c(log2FC_MS, log2FC_Olink),
               names_to =  c('.value', 'Platform'),
               names_sep = '_') %>% 
  ggplot.density(log2FC, Fill_var = Platform, median_text = F, median_line = F) +
  scale_fill_manual(values = TDP_palette[c(1,4)], aesthetics = c('color', 'fill')) +
  theme_publ() +
  labs(title = '', x = 'log2-fold change')


ggsave(filename = 'log2FC_density_noNA_overlappingProteins_sig.pdf',
       path = path,
       width = 8.7,
       height = 6.5,
       units = 'cm')

```



## Estimated concentration of DAPs


### Overlapping proteins - no missing values

```{r}

daa_res_conc_overlap <- daa_res_noNA_overlap$DAA_res |> 
  filter(adj.p < 0.05) %>% 
  inner_join(conc_data, by = join_by(UniProt == Uniprot))

ggplot.jitterbox(daa_res_conc_overlap, Platform, Blood.conc.ngmL) +
  scale_y_log10() +
  small_logticks('l') +
  stat_compare_means(size = 6/.pt, label.y = 5.5) +
  scale_color_manual(values = TDP_palette[c(1,4)], aesthetics = c('fill', 'color')) +
  labs(y = 'Concentration in blood (ng/mL)')

```



Overlapping proteins:

```{r}

dens_data <- daa_res_noNA_overlap$DAA_res  |> 
  filter(Significance == 'S') |>
  inner_join(conc_data, by = c('UniProt' = 'Uniprot'))

plot_data <- daa_res_noNA_overlap$DAA_res_wide  |> 
  filter(Significance != 'None') |> 
  inner_join(conc_data, by = c('UniProt' = 'Uniprot')) |> 
  mutate(Significance = droplevels(Significance))

stat_test <- wilcox_test(
  Blood.conc.ngmL ~ Significance, 
  data = plot_data) |> 
  add_xy_position(y.trans = log10)

set.seed(1)
plot_data |> 
  ggplot(aes(x = Significance, y = Blood.conc.ngmL)) +
  geom_jitter(aes(color = Significance), alpha = 0.8, width = 0.3, size = 0.5) +
  geom_boxplot(aes(fill = Significance), alpha = 0.5, outliers = F) +
  scale_y_log10(labels = log10_format(), breaks = 10^seq(-1,9,2)) +
  small_logticks('l') +
  labs(x = 'Significant in', 
       y = 'Concentration in blood (ng/mL)') +
  scale_fill_manual(values = daa_pal,
                    aesthetics = c('color', 'fill')) +
  theme_publ() +
  theme(legend.position = 'right',
        ggside.axis.line = element_blank(),
        ggside.axis.ticks = element_blank(),
        ggside.axis.text = element_blank(),
        ggside.panel.scale.y = 0.2) +
  geom_ysidedensity(data = dens_data,
                    mapping = aes(y = Blood.conc.ngmL, yfill = Platform, ycolor = Platform), 
                    alpha = 0.5) +
  scale_ycolor_manual(values = TDP_palette[c(1,4)], aesthetics = c('ycolor', 'yfill')) +
  guides(fill = 'none', color = 'none') +
  stat_pvalue_manual(stat_test, tip.length = 0.02, label.size = 6/.pt, step.increase = 0.1,
                     bracket.nudge.y = 0.5)


ggsave(filename = 'DAPs_by_conc_boxplot_noNA_overlappingProteins.pdf',
       path = path,
       width = 8.7,
       height = 6,
       units = 'cm')



```




