---
title: "Detected proteins, missing values, and proteome coverage"
author: "Noora Sissala"
date: "2025-03-19"
output: html_document
---


```{r setup}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(eulerr)
library(patchwork)
library(ggrepel)
library(scales)
library(rstatix)
library(ggpubr)
library(ggpp)

```

```{r}

# Load functions
source('code/utility_functions.R')
source('code/plotting_functions.R')

# Path for saving results
path <- 'results/proteome_coverage'

if (!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
}

# MS data
ms_data_all <- read_MS_data(withReplicateSamples = TRUE)

# MS data without replicates
ms_data <- read_MS_data()

# Olink Explore data

# All Olink assays
olink_data_all <- read_OlinkExplore_data(
  removeAllLOD = FALSE, 
  withReplicateAssays = TRUE)

# Detected Olink assays
olink_data <- read_OlinkExplore_data(withControlSamples = FALSE,
                                     withBelowLOD = TRUE,
                                     withQCWarning = TRUE,
                                     withReplicateAssays = TRUE,
                                     removeAllLOD = TRUE) |> 
  mutate(Below.LOD = NPX < LOD) # add column that indicates values below LOD

olink_data_wide <- olink_long_to_wide(olink_data, id_cols = c('UniProt', 'OlinkID'))

# Sample metadata
metadata <- read_metadata()
metadata_MS <- read_metadata(samples = 'MS')
metadata_all <- read_metadata(samples = 'MS', withReplicates = TRUE)

ms_data_overlap <- ms_data[colnames(ms_data) %in% colnames(olink_data_wide)]

# Protein metadata
protein_metadata <- read.csv('data/metadata/protein_metadata.csv') |> 
  filter(UniProt %in% ms_data$UniProt | OlinkID %in% olink_data$OlinkID)

# Load data from the Human Plasma Proteome Project (HPPP) (PeptideAtlas build 2023-04):
hppp <- read.csv('data/external_data/HPPP_build2023-04.csv')
hppp <- hppp$biosequence_accession

# Load data from the Human Protein Atlas
hpa_data <- read.delim('data/processed_data/HPA_v24_clean.txt')

# HPA plasma proteome: proteins with a known blood concentration or 
# annotated as secreted to blood
hpa_plasma <- hpa_data |> 
  filter(!is.na(Blood.conc.pgL) | Secretome.location == 'Secreted to blood')

# Reference plasma proteome
# HPPP + HPA plasma proteins
plasma_proteome <- union(hppp, hpa_plasma$Uniprot)
length(plasma_proteome)

```

## Cohort characteristics

### Number of samples

```{r}

# Calculate the number of proteins in each area of the venn diagram
n_ms <- length(setdiff(metadata_MS$Sample.ID, metadata$Sample.ID))
n_olink <- length(setdiff(metadata$Sample.ID, metadata_MS$Sample.ID))

n_overlap <- length(intersect(metadata$Sample.ID, metadata_MS$Sample.ID))

pdf(file = paste0(path, '/samples_venn.pdf'),
    width = 4*0.393701, height = 4*0.393701)
  # Plot the venn diagram
  p <- plot(euler(
    combinations = c('MS' = n_ms, 
                     'Olink' = n_olink, 
                     'MS&Olink' = n_overlap)), 
    fills = list(fill = TDP_palette[c(1,4)], alpha = 0.5), 
    edges = list(lwd = 2.5, col = TDP_palette[c(1,4)]), 
    quantities = list(fontsize = 6, font = 'regular', fontfamily = 'Helvetica'), 
    labels = list(fontsize = 7, font = 'regular', fontfamily = 'Helvetica')) 

  # Adjust plot to smaller size (make sure plot is not cut off when saving)
  p$vp$width <- unit(0.95, "npc")
  p$vp$height <- unit(0.95, "npc")
  
  grid::grid.draw(p)

dev.off()



```


### Age and sex distribution


```{r}

tryCatch({
  
  participant_metadata <- read.delim('data/metadata/participant_metadata.txt')
  
  age_distr <- ggplot.density(
    participant_metadata, Age, fill = '#CFD8F2', color = '#A8BAE6',
    median_text = FALSE,
    median_line_params = list(lwd = 0.3)) +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=expansion(c(0,0.02))) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_line(color = 'black'),
        panel.background = element_rect(fill='transparent'),
        plot.background = element_rect(fill='transparent', color=NA),
        text = element_text(size = 8))

  label_df <- data.frame(Sex = c('Female', 'Male'),
                         x = c(1, 1),
                         y = c(0.8, 0.3),
                         N = as.numeric(table(participant_metadata$Sex))) |> 
    mutate(label = paste0(Sex, '\n', N))
  
  sex_distr <- ggplot(participant_metadata, aes(x = 1, fill = Sex)) +
    geom_bar(position = 'fill', show.legend = F) +
    geom_text(data = label_df, aes(label = label, x = x, y = y),
              size = 8/.pt) +
    labs(x = 'Sex') +
    theme_void() +
    scale_y_continuous(expand=expansion(c(0,0.02))) +
    theme(axis.title.x = element_text(size = 8)) +
    scale_fill_manual(values = as.character(TDP_palette[c(14,11)]))
  
  
  p <- age_distr + sex_distr + plot_layout(widths = c(2.5,1)) & 
    theme(panel.background = element_rect(fill='transparent', color = NA),
          plot.background = element_rect(fill='transparent', color=NA))
  
  
  ggsave(p,
         filename = 'cohort_characteristics.png',
         path = path,
         width = 4.25,
         height = 3.25,
         units = 'cm',
         bg = 'transparent')
  
},

error = function(e) {
  message('Age and sex not publicly available')
}
)


```


## Detected proteins and MS-Olink overlap


Table of detected proteins:

```{r}

missing_MS <- enframe(
  setNames(rowSums(is.na(ms_data_overlap[-1])), ms_data_overlap$UniProt), 
  name = 'UniProt', value = 'N.Missing.MS') |> 
  mutate(Perc.Missing.MS = N.Missing.MS/(ncol(ms_data_overlap)-1)*100)

ms_protein_metadata <- read.csv('data/metadata/protein_metadata_MS.csv',
                           row.names = 1) |> 
  dplyr::select(1:3) |> 
  inner_join(missing_MS[c('UniProt', 'Perc.Missing.MS')], by = 'UniProt')

missing_Olink <- olink_data_all |> 
  mutate(Below.LOD = NPX < LOD) |> 
  group_by(UniProt, OlinkID) |>
  summarize(N.Missing.Olink = sum(Below.LOD),
            Perc.Missing.Olink = N.Missing.Olink/n()*100)

olink_protein_metadata <- olink_data_all|> 
  distinct(OlinkID, UniProt, Assay, Panel, Panel_Lot_Nr, PlateID) |> 
  left_join(missing_Olink[c('UniProt', 'OlinkID', 'Perc.Missing.Olink')], 
            by = c('UniProt', 'OlinkID'))

protein_table <- full_join(ms_protein_metadata, olink_protein_metadata, by = 'UniProt') |> 
  mutate(Analyzed.MS = !is.na(Gene.Name),
         Analyzed.Olink = !is.na(Assay),
         Detected.MS = Analyzed.MS & Perc.Missing.MS < 100,
         Detected.Olink = Analyzed.Olink & Perc.Missing.Olink < 100,
         Overlapping = Detected.MS & Detected.Olink,
         Detected.Platform = case_when(Detected.MS & Detected.Olink ~ 'Both',
                                       Detected.MS & !Detected.Olink ~ 'MS',
                                       Detected.Olink & !Detected.MS ~ 'Olink',
                                       TRUE ~ 'None'),
         In.HPA = UniProt %in% hpa_data$Uniprot, # Is protein in the HPA
         In.HPA.Plasma = UniProt %in% hpa_plasma$Uniprot, # Does protein have a blood concentration in HPA or is it classified as secreted to blood
         In.HPPP = UniProt %in% hppp, # Is protein in the Plasma section of the PeptideAtlas (Human Plasma Proteome Project)
         In.Ref.Plasma.Proteome = UniProt %in% plasma_proteome) |> # Is protein in the reference plasma proteome
  dplyr::select(UniProt, Gene.Name, Description, # MS protein information
                OlinkID, Assay, Panel, Panel_Lot_Nr, PlateID, # Olink protein information
                Analyzed.MS, Detected.MS, Perc.Missing.MS, # MS detection information
                Analyzed.Olink, Detected.Olink, Perc.Missing.Olink, # Olink detection information
                everything())

write.csv(protein_table, file.path(path, 'detected_proteins.csv'),
          row.names = FALSE)

```

```{r}

# Detected proteins with no replicate Olink assays
detected_proteins <- protein_table |> 
  group_by(UniProt) |> 
  summarize(Detected.MS = any(Detected.MS),
            Detected.Olink = any(Detected.Olink),
            In.HPA = any(In.HPA),
            In.HPA.Plasma = any(In.HPA.Plasma),
            In.HPPP = any(In.HPPP),
            In.Ref.Plasma.Proteome = any(In.Ref.Plasma.Proteome)) |> 
  dplyr::rename(MS = Detected.MS, Olink = Detected.Olink)

# Vectors of detected proteins
olink_proteins <- unique(olink_data$UniProt)
ms_proteins <- unique(ms_data$UniProt)
all_proteins <- union(olink_proteins, ms_proteins)
overlap <- intersect(olink_proteins, ms_proteins)

```


Venn diagram of detected proteins:

```{r}

pdf(file = paste0(path, '/venn_MS_Olink.pdf'),
    width = 5*0.393701, height = 5*0.393701)

  p <- plot(euler(
    combinations = detected_proteins[c('MS', 'Olink')]),
    fills = list(fill = TDP_palette[c(1,4)], alpha = 0.5),
    edges = list(lwd = 2.5, col = TDP_palette[c(1,4)]), 
    quantities = list(fontsize = 7, font = 'regular', fontfamily = 'Helvetica'), 
    labels = list(fontsize = 8, font = 'regular', fontfamily = 'Helvetica'))

  # Adjust plot to smaller size (make sure plot is not cut off when saving)
  p$vp$width <- unit(0.95, "npc")
  p$vp$height <- unit(0.95, "npc")
  
  grid::grid.draw(p)

dev.off()


```

```{r}

p

```


## Overlap with reference plasma proteome

The reference plasma proteome was compiled from proteins in the Plasma section of the PeptideAtlas + proteins with a known blood concentration or secretome location = "Secreted to blood" in the Human Protein Atlas (HPA).


Venn diagram of detected proteins and the reference plasma proteome:

```{r}

all_proteins_df <- data.frame(UniProt = plasma_proteome) |> 
  mutate(Plasma.proteome = TRUE) |> 
  full_join(detected_proteins[c('UniProt', 'MS', 'Olink')], by = 'UniProt') |> 
  mutate(MS = ifelse(is.na(MS), FALSE, MS),
         Olink = ifelse(is.na(Olink), FALSE, Olink),
         Plasma.proteome = ifelse(is.na(Plasma.proteome), FALSE, Plasma.proteome))

pdf(file = paste0(path, '/venn_MS_Olink_plasmaProteome.pdf'),
    width = 4*0.393701, height = 5*0.393701)

  p <- plot(euler(
    combinations = setNames(all_proteins_df[c('MS', 'Olink', 'Plasma.proteome')],
                            c('MS', 'Olink', 'Plasma proteome'))),
    fills = list(fill = TDP_palette[c(1,4,7)], alpha = 0.5), 
    edges = list(lwd = 2.5, col = TDP_palette[c(1,4,7)]),
    quantities = list(fontsize = 7, font = 'regular', fontfamily = 'Helvetica'), 
    labels = list(fontsize = 8, font = 'regular', fontfamily = 'Helvetica'))

  # Adjust plot to smaller size
  p$vp$width <- unit(0.95, "npc")
  p$vp$height <- unit(0.95, "npc")
  
  grid::grid.draw(p)

dev.off()


```

```{r}

p

```


## Plasma proteome coverage


Bar plot of plasma proteome coverage:

```{r}

# Calculate proportion of proteins in MS and/or Olink that are in the reference proteome
ms_prop <- length(intersect(ms_proteins, plasma_proteome)) / length(plasma_proteome)*100
olink_prop <- length(intersect(olink_proteins, plasma_proteome)) / length(plasma_proteome)*100
ms_olink_prop <- length(intersect(all_proteins, plasma_proteome)) / length(plasma_proteome)*100

plot_data <- data.frame(
  Protein_Set = factor(c(rep("MS", 2), rep("Olink", 2), rep("Both", 2)),
                       levels = c('MS', 'Olink', 'Both')),
  Source = factor(c('MS', 'Plasma proteome', 'Olink', 'Plasma proteome', 'Both', 'Plasma proteome'),
                  levels = c('Plasma proteome', 'Both', 'Olink', 'MS')),
  Proportion = c(ms_prop, 100 - ms_prop, 
                 olink_prop, 100 - olink_prop, 
                 ms_olink_prop, 100 - ms_olink_prop)) |> 
  mutate(Label = paste0(round(Proportion), '%'),
         Label = ifelse(Source == 'Plasma proteome', NA, Label))


```


```{r}

plot_data |> 
  ggplot(aes(fill = Source, alpha = Source, y = Proportion, x = Protein_Set)) +
  geom_col() +
  geom_text(aes(label = Label),
            position = position_stack(vjust = 0.5), size = 5/.pt) +
  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
  theme_publ() +
  labs(x = '', y = '') +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 5),
        legend.position = 'none') +
  scale_fill_manual(values = as.character(TDP_palette[c(7,13,4,1)])) +
  scale_alpha_manual(values = c(0.8,1,1,1))
    
    

```

```{r}

ggsave(filename = 'plasma_proteome_coverage.pdf',
       path = path,
       width = 2.2,
       height = 1.8,
       units = 'cm')


```


## Overlap by Olink panel

```{r}

# Get the total number of protein assays per panel
proteins_per_panel <- olink_data_all |>
    group_by(Panel) |>
    summarise(total.proteins = length(unique(UniProt)))

# Calculate number of protein assays in each panel that were detected with both methods
overlap_per_panel <- olink_data %>% 
  filter(UniProt %in% overlap) %>% 
  group_by(Panel) %>% 
  summarize(N = n_distinct(UniProt)) %>%
  arrange(Panel) %>%
  inner_join(proteins_per_panel, by = 'Panel') %>%
  mutate(Perc = N/total.proteins*100,
         Panel = as.factor(Panel))

ggplot.barplot(overlap_per_panel, X_var = N, Y_var = Panel, Fill_var = Panel,
               title = 'Overlap by Olink Explore panel',
               show.legend = F) +
  scale_y_discrete(limits = rev) +
  geom_text(aes(x = N, y = Panel, label = paste0(round(Perc, 1), '%')),
            size = 6/.pt, hjust = -0.1) +
  expand_limits(x = c(0,300)) +
  scale_fill_manual(values = panel_pal) +
  theme(legend.position = 'none') +
  labs(x = 'Number of proteins') +
  theme_publ() +
  theme(plot.title = element_text(size = 7),
        axis.title.y = element_blank())


```

```{r, include=FALSE}

ggsave(filename = 'overlap_per_panel.pdf',
       path = path,
       width = 5.5,
       height = 6,
       units = 'cm')

```


## Missing values and detected proteins


Check whether replicate Olink assays have values <LOD

```{r}

replicate_assays <- olink_data |> 
  group_by(UniProt) |>
  summarise(N = n_distinct(OlinkID)) |> 
  filter(N > 1) |> 
  pull(UniProt)

missing_Olink |>  filter(UniProt %in% replicate_assays, Perc.Missing.Olink > 0)

```

No replicate assays have values <LOD -> include one assay per protein in the remaining analyses.


Summary of missing values and detected proteins:

```{r}

# Select one Olink assay per protein
set.seed(1)
missing_Olink <- missing_Olink |> 
  group_by(UniProt) |>
  sample_n(1)

labels <- c('0', '(0-25]', '(25-50]', '(50-75]', '(75-100]')

# Divide missing values into intervals and calculate cumulative sums
# and proportions
plot_data <- bind_rows('MS' = missing_MS |> 
                         dplyr::rename(N.Missing = N.Missing.MS, 
                                       Perc.Missing = Perc.Missing.MS), 
                       'Olink' = missing_Olink |> 
                         dplyr::rename(N.Missing = N.Missing.Olink, 
                                       Perc.Missing = Perc.Missing.Olink),
                       .id = 'Platform') |> 
  mutate(Perc.missing.interval = as.character(cut(
    Perc.Missing, breaks = seq(-25, 100, 25), labels = labels, include.lowest = F, right = T)),
    Perc.missing.interval = factor(Perc.missing.interval, levels = labels)) |>
  group_by(Platform, Perc.missing.interval) %>%
  summarize(N = n()) %>% 
  mutate(Cumulative = cumsum(N),
         Perc = N/sum(N)*100,
         Cumulative.perc = Cumulative/sum(N)*100)

# Plot missing values and detected proteins
ggplot(plot_data, aes(x = Perc.missing.interval, y = Perc, fill = Platform, color = Platform)) + 
  geom_col(position = position_dodge(width = 0.8), width = 0.73) +
  geom_point(aes(y = Cumulative.perc), size = 1) +
  geom_line(aes(y = Cumulative.perc, group = Platform),
            linewidth = 0.8) +
  geom_label_repel(aes(y = Cumulative.perc, label = Cumulative), 
             fill = 'white', show.legend = F, size = 6/.pt, direction = 'y',
             nudge_y = 5, label.padding = 0.1) +
  labs(x = 'Proportion of missing values (%)', y = 'Proportion of proteins (%)', 
       color = 'Platform', fill = 'Platform', linetype = '') +
  theme_publ() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 100, 20),
                     limits = c(0, 110)) +
  scale_color_manual(values = TDP_palette[c(1,4)], aesthetics = c('color', 'fill')) +
  scale_linetype_manual(values = c('Cumulative sum' = 'solid')) 

```

```{r}

ggsave(filename = 'detectedProteins_and_NAs_combined.pdf',
       path = path,
       width = 6.5,
       height = 6.5,
       unit = 'cm',
       device = 'pdf')

```


Proteins with any missing values

```{r}

missing_MS |> 
  ungroup() |> 
  mutate(Any.missing = Perc.Missing.MS > 0) |>
  summarize(Perc.any.missing = sum(Any.missing)/n()*100,
            N.any.missing = sum(Any.missing),
            N.no.missing = sum(!Any.missing),
            N.total = n())

missing_Olink |>
  ungroup() |>
  mutate(Any.missing = Perc.Missing.Olink > 0) |>
  summarize(Perc.any.missing = sum(Any.missing)/n()*100,
            N.any.missing = sum(Any.missing),
            N.no.missing = sum(!Any.missing),
            N.total = n())

```



## Missing values per sample (MS)

```{r}

NA_per_sample <- ms_data_all %>%
  pivot_longer(cols = -UniProt,
               values_to = 'Value',
               names_to = 'Sample.ID') %>%
  group_by(Sample.ID, .drop = FALSE) %>%
  summarize(N.missing = sum(is.na(Value)),
            Perc.missing = sum(is.na(Value))/n()*100,
            Total = n()) %>%
  ungroup() |> 
  left_join(metadata_all, by = 'Sample.ID') |> 
  mutate(TMT.set = factor(TMT.set, 
                          levels = c('setA', 'setB', 'setC', 'setD', 'setE',
                                     'setF', 'setG', 'setH'))) |> 
  arrange(TMT.set)
                                              


ggplot(NA_per_sample, aes(x = reorder(Sample.ID, Perc.missing), y = Perc.missing,
                          fill = TMT.set)) + 
  geom_col() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_publ() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = 'Sample', y = 'Proportion of missing values (%)', fill = 'TMT set') +
  scale_fill_manual(values = desaturate(lighten(hue_pal()(8), 0.3), 0.2))
  

```

```{r}

ggsave(filename = 'NAs_per_sample.pdf',
       path = path,
       width = 12.5,
       height = 8.5,
       units = 'cm',
       device = 'pdf')

```


## Missing values vs. concentration in blood

```{r}

conc_data <- hpa_plasma %>% 
  filter(!is.na(Blood.conc.ngmL)) %>% 
  dplyr::select(Uniprot, Blood.conc.ngmL) |> 
  distinct()

```

```{r}

labels <- c('0', '(0-25]', '(25-50]', '(50-75]', '(75-100]')

plot_data <- bind_rows(
  'MS' = missing_MS |> 
    dplyr::rename(N.Missing = N.Missing.MS, Perc.Missing = Perc.Missing.MS), 
  'Olink' = missing_Olink |>
    dplyr::rename(N.Missing = N.Missing.Olink, Perc.Missing = Perc.Missing.Olink), 
  .id = 'Platform') |>
  inner_join(conc_data, by = join_by(UniProt == Uniprot)) |> 
  mutate(Interval = cut(
    Perc.Missing, breaks = seq(-25,100,25), labels = labels, 
    right = TRUE, include.lowest = FALSE))

# Calculate p-values
stat_df <- plot_data |> 
  group_by(Interval) |>
  wilcox_test(Blood.conc.ngmL ~ Platform) |> 
  adjust_pvalue(method = 'fdr') |>
  add_significance('p.adj') |> 
  add_xy_position(x = 'Interval', y.trans = log10)

conc_plot <- plot_data |> 
  ggplot.jitterbox(Interval, Blood.conc.ngmL, Color_var = Platform) +
  scale_y_continuous(trans = 'log10', labels = log10_format(),
                     breaks = 10^seq(-2,9,2)) +
  labs(x = 'Proportion of missing values (%)', y = 'Concentration in blood (ng/mL)') +
  scale_color_manual(values = TDP_palette[c(1,4)], aesthetics = c('color', 'fill')) +
  stat_pvalue_manual(stat_df, inherit.aes = F, y.position = 8.5,
                     tip.length = 0.02, size = 6/.pt)

conc_plot

```

```{r}

ggsave(conc_plot,
       filename = 'NAs_vs_conc_boxplot.pdf',
       path = path,
       width = 8.3, 
       height = 6.5,
       units = 'cm',
       device = 'pdf')

```


## Concentration of detected proteins

### All detected proteins

```{r}

conc_data_detected <- conc_data %>% 
  mutate(Platform = case_when(
    Uniprot %in% ms_proteins & ! Uniprot %in% olink_proteins ~ 'MS',
    Uniprot %in% olink_proteins & ! Uniprot %in% ms_proteins ~ 'Olink',
    Uniprot %in% ms_proteins & Uniprot %in% olink_proteins ~ 'MS,Olink')) %>% 
  separate_longer_delim(cols = Platform, ',') |> 
  mutate(Platform = factor(Platform, levels = c('MS', 'Olink'))) |> 
  filter(!is.na(Platform))

```


```{r}

# Calculate p-value for difference in concentration between platforms
stat_df <- conc_data_detected %>% 
  group_by(Platform) %>% 
  summarize(Median = round(median(Blood.conc.ngmL, na.rm = T), 2),
            Q1 = round(quantile(Blood.conc.ngmL, 0.25, na.rm = T), 2),
            Q3 = round(quantile(Blood.conc.ngmL, 0.75, na.rm = T), 2)) |>  
  mutate(y = c(0.43,0.45),
         npcy = c(0.9,0.95))

# P-value label for plot
pval <- wilcox.test(Blood.conc.ngmL ~ Platform, data = conc_data_detected)$p.value
pval_label = format_pval_label(pval)

# Concentration density plot
p1 <- ggplot(conc_data_detected, 
             aes(x = Blood.conc.ngmL, color = Platform, fill = Platform)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(trans = 'log10',
                     breaks = 10^(seq(-2,8,2)),
                     expand = expansion(c(0,0)),
                     labels = trans_format('log10', math_format(10^.x))) +
  scale_y_continuous(expand = expansion(c(0, 0.05)),
                     limits = c(0,0.45)) +
  theme_publ() +
  labs(x = 'Concentration in blood (ng/mL)', y = 'Density',
       title = 'All detected proteins') +
  scale_color_manual(values = TDP_palette[c(1,4)],
                     aesthetics = c('color', 'fill')) +
  # add point and line for median and IQR
  geom_point(
    data = stat_df,
    mapping = aes(x = Median, y = y, color = Platform),
    size = 1.5) +
  geom_segment(
    data = stat_df,
    mapping = aes(x = Q1, xend = Q3, color = Platform, y = y),
    linewidth = 0.5) +
  # add p.value
  geom_text_npc(data = data.frame(
    npcx = 'right', npcy = 0.95, label = pval_label), 
    mapping = aes(npcx = npcx, npcy = npcy, 
                  label = label),
    size = 6/.pt, hjust = 1, parse = TRUE)

p1

```


### Uniquely detected proteins


```{r}

conc_data_detected <- conc_data_detected %>% 
  mutate(All.detected.proteins = TRUE,
         Detected.one.method.only = Uniprot %in% setdiff(all_proteins, overlap))

unique_conc_data <- filter(conc_data_detected, Detected.one.method.only)

```


```{r}

# Calculate p-value for difference in concentration between platforms
stat_df_unique <- unique_conc_data %>% 
  group_by(Platform) %>% 
  summarize(Median = round(median(Blood.conc.ngmL, na.rm = T), 2),
            Q1 = round(quantile(Blood.conc.ngmL, 0.25, na.rm = T), 2),
            Q3 = round(quantile(Blood.conc.ngmL, 0.75, na.rm = T), 2)) |>  
  mutate(y = c(0.43,0.45),
         npcy = c(0.9,0.95))

# P-value label for plot
pval_unique <- wilcox.test(Blood.conc.ngmL ~ Platform, data = unique_conc_data)$p.value
pval_label_unique = format_pval_label(pval_unique)

# Concentration density plot
p2 <- ggplot(unique_conc_data, aes(x = Blood.conc.ngmL, color = Platform, fill = Platform)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(trans = 'log10',
                     breaks = 10^(seq(-2,7,2)),
                     expand = expansion(c(0,0)),
                     labels = trans_format('log10', math_format(10^.x))) +
  scale_y_continuous(expand = expansion(c(0, 0.05)),
                     limits = c(0,0.45)) +
  theme_publ() +
  labs(x = 'Concentration in blood (ng/mL)', y = 'Density',
       title = 'Detected with one method only') +
  scale_color_manual(values = TDP_palette[c(1,4)],
                     aesthetics = c('color', 'fill')) +
  # add point and line for median and IQR
  geom_point(
    data = stat_df_unique,
    mapping = aes(x = Median, y = y, color = Platform),
    size = 1.5) +
  geom_segment(
    data = stat_df_unique,
    mapping = aes(x = Q1, xend = Q3, color = Platform, y = y),
    linewidth = 0.5) +
  # add p.value
  geom_text_npc(data = data.frame(
    npcx = 'right', npcy = 0.95, label = pval_label_unique), 
    mapping = aes(npcx = npcx, npcy = npcy, 
                  label = label),
    size = 6/.pt, hjust = 1, parse = TRUE)

p2


```

```{r}

# Combine plots
guide_area() +  p1 + p2 + 
  plot_layout(design = 'AA\nBC', guides = 'collect', axis_titles = 'collect', 
              heights = c(0.4, 10), nrow = 2, ncol = 1)

ggsave(filename = 'HPA_bloodConc_density_plots.pdf',
       path = path,
       width = 12,
       height = 6.5,
       units = 'cm')

```

## Plasma proteome coverage by concentration interval

```{r}

# Which proteins were detected by each platform
conc_prop <- conc_data %>% 
  mutate(Platform = case_when(
    Uniprot %in% ms_proteins & ! Uniprot %in% olink_proteins ~ 'MS',
    Uniprot %in% olink_proteins & ! Uniprot %in% ms_proteins ~ 'Olink',
    Uniprot %in% ms_proteins & Uniprot %in% olink_proteins ~ 'Both',
    !(Uniprot %in% ms_proteins | Uniprot %in% olink_proteins) ~ 'Not detected')) %>% 
  mutate(Platform = factor(Platform, levels = c('Both', 'MS', 'Olink', 'Not detected')))
               
conc_prop |>     
ggplot.histogram(Blood.conc.ngmL, Fill_var = Platform, breaks = 10^(seq(-3, 8, 0.5)),
                   position = position_fill(reverse = T), median_line = F,
                   median_text = F, alpha = 0.9) +
  scale_x_continuous(trans = 'log10', expand = expansion(c(0,0.01)), 
                     breaks = 10^(seq(-3, 8, 1)),
                     labels = log10_format()) +
  scale_y_continuous(expand = expansion(c(0,0.01))) +
  labs(x = 'Concentration in blood (ng/mL)',
       y = 'Proportion of plasma proteins') +
  scale_color_manual(values = c('Not detected' = TDP_palette[[7]],
                                'Both' = TDP_palette[[13]],
                                TDP_palette[c(1,4)]),
                     aesthetics = c('color', 'fill')) +
  theme_publ()

```

```{r}

ggsave(filename = 'HPA_bloodConc_plasmaProteomeCoverage.pdf',
       path = path,
       width = 8.3,
       height = 6.5,
       units = 'cm')


```



## Proteins with max 50% missing values

Note: proportion of missing values for MS was calculated based on overlapping samples (N = 88), not all 120.

### Detected proteins and MS-Olink overlap

```{r}

protein_table_max50NA <- protein_table |> 
  filter(Perc.Missing.MS <= 50 | Perc.Missing.Olink <= 50) |> 
  mutate(Detected.MS = Perc.Missing.MS <= 50,
         Detected.MS = ifelse(is.na(Detected.MS), FALSE, Detected.MS),
         Detected.Olink = Perc.Missing.Olink <= 50,
         Detected.Olink = ifelse(is.na(Detected.Olink), FALSE, Detected.Olink))

detected_proteins_max50NA <- protein_table_max50NA |> 
  group_by(UniProt) |> 
  summarize(Detected.MS = any(Detected.MS),
            Detected.Olink = any(Detected.Olink),
            In.HPA = any(In.HPA),
            In.HPA.Plasma = any(In.HPA.Plasma),
            In.HPPP = any(In.HPPP),
            In.Ref.Plasma.Proteome = any(In.Ref.Plasma.Proteome)) |> 
  dplyr::rename(MS = Detected.MS, Olink = Detected.Olink)


ms_proteins_filtered <- missing_MS |> 
  filter(Perc.Missing.MS <= 50) |> 
  pull(UniProt)

olink_proteins_filtered <- missing_Olink |> 
  filter(Perc.Missing.Olink <= 50) |> 
  pull(UniProt)

overlap_filtered <- unique(intersect(ms_proteins_filtered, olink_proteins_filtered))
all_proteins_filtered <- unique(union(ms_proteins_filtered, olink_proteins_filtered))

```


Venn diagram of detected proteins with max 50% missing values:

```{r}


pdf(file = paste0(path, '/venn_MS_Olink_max50percNA.pdf'),
    width = 4*0.393701, height = 4*0.393701)
  p <- plot(euler(
    combinations = detected_proteins_max50NA[c('MS', 'Olink')]),
    fills = list(fill = TDP_palette[c(1,4)], alpha = 0.5), 
    edges = list(lwd = 2.5, col = TDP_palette[c(1,4)]), 
    quantities = list(fontsize = 7, font = 'regular', fontfamily = 'Helvetica'), 
    labels = list(fontsize = 8, font = 'regular', fontfamily = 'Helvetica')) 

  p$vp$width <- unit(0.95, "npc")
  p$vp$height <- unit(0.95, "npc")
  
  grid::grid.draw(p)
  
dev.off()

```

```{r}

p

```



### Overlap with reference plasma proteome

Venn diagram of detected proteins with max 50% missing values and the reference plasma proteome:

```{r}

all_proteins_max50NA_df <- data.frame(UniProt = plasma_proteome) |> 
  mutate(Plasma.proteome = TRUE) |> 
  full_join(detected_proteins_max50NA[c('UniProt', 'MS', 'Olink')], by = 'UniProt') |> 
  mutate(MS = ifelse(is.na(MS), FALSE, MS),
         Olink = ifelse(is.na(Olink), FALSE, Olink),
         Plasma.proteome = ifelse(is.na(Plasma.proteome), FALSE, Plasma.proteome))

pdf(file = paste0(path, '/venn_MS_Olink_plasmaProteome_max50percNA.pdf'),
    width = 4*0.393701, height = 5*0.393701)

  p <- plot(euler(
    combinations = setNames(all_proteins_max50NA_df[c('MS', 'Olink', 'Plasma.proteome')],
                            c('MS', 'Olink', 'Plasma proteome'))),
    fills = list(fill = TDP_palette[c(1,4,7)], alpha = 0.5), 
    edges = list(lwd = 2.5, col = TDP_palette[c(1,4,7)]),
    quantities = list(fontsize = 7, font = 'regular', fontfamily = 'Helvetica'), 
    labels = list(fontsize = 8, font = 'regular', fontfamily = 'Helvetica'))

  p$vp$width <- unit(0.95, "npc")
  p$vp$height <- unit(0.95, "npc")
  
  grid::grid.draw(p)
  
dev.off()


```

```{r}

p

```


### Plasma proteome coverage


```{r}

ms_prop <- length(intersect(ms_proteins_filtered, plasma_proteome)) / length(plasma_proteome)*100
olink_prop <- length(intersect(olink_proteins_filtered, plasma_proteome)) / length(plasma_proteome)*100
ms_olink_prop <- length(intersect(all_proteins_filtered, plasma_proteome)) / length(plasma_proteome)*100

plot_data <- data.frame(
  Protein_Set = factor(c(rep("MS", 2), rep("Olink", 2), rep("Both", 2)),
                       levels = c('MS', 'Olink', 'Both')),
  Source = factor(c('MS', 'Plasma proteome', 'Olink', 'Plasma proteome', 'Both', 'Plasma proteome'),
                  levels = c('Plasma proteome', 'Both', 'Olink', 'MS')),
  Proportion = c(ms_prop, 100 - ms_prop, 
                 olink_prop, 100 - olink_prop, 
                 ms_olink_prop, 100 - ms_olink_prop)) |> 
  mutate(Label = paste0(round(Proportion), '%'),
         Label = ifelse(Source == 'Plasma proteome', NA, Label))


```


```{r}

plot_data |> 
  ggplot(aes(fill = Source, alpha = Source, y = Proportion, x = Protein_Set)) +
  geom_col() +
  geom_text(aes(label = Label),
            position = position_stack(vjust = 0.5), size = 5/.pt) +
  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
  theme_publ() +
  labs(x = '', y = '') +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 5),
        legend.position = 'none') +
  scale_fill_manual(values = as.character(TDP_palette[c(7,13,4,1)])) +
  scale_alpha_manual(values = c(0.8,1,1,1))
    
    

```

```{r}

ggsave(filename = 'plasma_proteome_coverage_max50percNA.pdf',
       path = path,
       width = 2.2,
       height = 1.8,
       units = 'cm')


```


### Overlap per Olink panel

```{r}

overlap_per_panel_max50NA <- olink_data_all %>% 
  filter(UniProt %in% overlap_filtered) %>% 
  group_by(Panel) %>% 
  summarize(N = n_distinct(UniProt)) %>%
  arrange(Panel) %>%
  inner_join(proteins_per_panel, by = 'Panel') %>%
  mutate(Perc = N/total.proteins*100,
         Panel = as.factor(Panel))

ggplot.barplot(overlap_per_panel_max50NA, X_var = N, Y_var = Panel, Fill_var = Panel,
               title = 'Overlap by Olink Explore panel',
               show.legend = F) +
  scale_y_discrete(limits = rev) +
  geom_text(aes(x = N, y = Panel, label = paste0(round(Perc, 1), '%')),
            size = 6/.pt, hjust = -0.1) +
  expand_limits(x = c(0,300)) +
  scale_fill_manual(values = panel_pal) +
  theme(legend.position = 'none') +
  labs(x = 'Number of proteins') +
  theme_publ()

```

```{r, include=FALSE}

ggsave(filename = 'overlap_per_panel_max50percNA.pdf',
       path = path,
       width = 6.5,
       height = 6,
       units = 'cm')

```


### Concentration of detected proteins


All proteins detected with max 50% missing values:

```{r}

conc_data_detected_max50NA <- conc_data %>%
  mutate(Platform = case_when(
    Uniprot %in% ms_proteins_filtered & !Uniprot %in% olink_proteins_filtered ~ 'MS',
    Uniprot %in% olink_proteins_filtered & !Uniprot %in% ms_proteins_filtered ~ 'Olink',
    Uniprot %in% ms_proteins_filtered & Uniprot %in% olink_proteins_filtered ~ 'MS,Olink')) %>% 
  separate_longer_delim(cols = Platform, ',') |> 
  filter(!is.na(Platform))

```

```{r}

# Calculate p-value for difference in concentration between platforms
stat_df_max50NA <- conc_data_detected_max50NA %>% 
  group_by(Platform) %>% 
  summarize(Median = round(median(Blood.conc.ngmL, na.rm = T), 2),
            Q1 = round(quantile(Blood.conc.ngmL, 0.25, na.rm = T), 2),
            Q3 = round(quantile(Blood.conc.ngmL, 0.75, na.rm = T), 2)) |>  
  mutate(y = c(0.4,0.42),
         npcy = c(0.9,0.95))

# P-value label for plot
pval_max50NA <- wilcox.test(Blood.conc.ngmL ~ Platform, data = conc_data_detected_max50NA)$p.value
pval_label_max50NA <- format_pval_label(pval_max50NA)

# Concentration density plot
p1 <- ggplot(conc_data_detected_max50NA, 
             aes(x = Blood.conc.ngmL, color = Platform, fill = Platform)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(trans = 'log10',
                     breaks = 10^(seq(-2,7,2)),
                     expand = expansion(c(0,0)),
                     labels = log10_format()) +
  scale_y_continuous(expand = expansion(c(0, 0.05)),
                     limits = c(0,0.43)) +
  theme_publ() +
  labs(x = 'Concentration in blood (ng/mL)', y = 'Density',
       title = 'All detected proteins') +
  scale_color_manual(values = TDP_palette[c(1,4)],
                     aesthetics = c('color', 'fill')) +
  # add point and line for median and IQR
  geom_point(
    data = stat_df_max50NA,
    mapping = aes(x = Median, y = y, color = Platform),
    size = 1.5) +
  geom_segment(
    data = stat_df_max50NA,
    mapping = aes(x = Q1, xend = Q3, color = Platform, y = y),
    linewidth = 0.5) +
  # add p.value
  geom_text_npc(data = data.frame(
    npcx = 'right', npcy = 0.95, label = pval_label_max50NA), 
    mapping = aes(npcx = npcx, npcy = npcy, 
        label = label),
    size = 6/.pt, hjust = 1, parse = TRUE)

p1

```


Uniquely detected proteins with max 50% missing values:


```{r}

conc_data_detected_max50NA <- conc_data_detected_max50NA %>% 
  mutate(All.detected.proteins = TRUE,
         Detected.one.method.only = Uniprot %in% setdiff(all_proteins_filtered, overlap_filtered))

unique_conc_data_max50NA <- filter(conc_data_detected_max50NA, Detected.one.method.only)

```

```{r}

# Calculate p-value for difference in concentration between platforms
stat_df_max50NA_unique <- unique_conc_data_max50NA %>% 
  group_by(Platform) %>% 
  summarize(Median = round(median(Blood.conc.ngmL, na.rm = T), 2),
            Q1 = round(quantile(Blood.conc.ngmL, 0.25, na.rm = T), 2),
            Q3 = round(quantile(Blood.conc.ngmL, 0.75, na.rm = T), 2)) |>  
  mutate(y = c(0.4,0.42),
         npcy = c(0.9,0.95))

# P-value label for plot
pval_max50NA_unique <- wilcox.test(Blood.conc.ngmL ~ Platform, data = unique_conc_data_max50NA)$p.value
pval_label_max50NA_unique <- format_pval_label(pval_max50NA_unique)

# Concentration density plot
p2 <- ggplot(unique_conc_data_max50NA, aes(x = Blood.conc.ngmL, color = Platform, fill = Platform)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(trans = 'log10',
                     breaks = 10^(seq(-2,7,2)),
                     expand = expansion(c(0,0)),
                     labels = log10_format()) +
  scale_y_continuous(expand = expansion(c(0, 0.05)),
                     limits = c(0,0.43)) +
  theme_publ() +
  labs(x = 'Concentration in blood (ng/mL)', y = 'Density',
       title = 'Detected with one method only') +
  scale_color_manual(values = TDP_palette[c(1,4)],
                     aesthetics = c('color', 'fill')) +
  # add point and line for median and IQR
  geom_point(
    data = stat_df_max50NA_unique,
    mapping = aes(x = Median, y = y, color = Platform),
    size = 1.5) +
  geom_segment(
    data = stat_df_max50NA_unique,
    mapping = aes(x = Q1, xend = Q3, color = Platform, y = y),
    linewidth = 0.5) +
  # add p.value
  geom_text_npc(data = data.frame(
    npcx = 'right', npcy = 0.95, label = pval_label_max50NA_unique), 
    mapping = aes(npcx = npcx, npcy = npcy, 
                  label = label),
    size = 6/.pt, hjust = 1, parse = TRUE)

p2


```

```{r}

## Combine plots
guide_area() +  p1 + p2 +  # put legend above plot titles
  plot_layout(design = 'AA\nBC', guides = 'collect', axis_titles = 'collect', 
              heights = c(0.4, 10), nrow = 2, ncol = 1)

ggsave(filename = 'HPA_bloodConc_density_plots_max50NA.pdf',
       path = path,
       width = 11.3,
       height = 6.5,
       units = 'cm')

```

### Plasma proteome coverage by concentration interval

```{r}

# Which proteins were detected by each platform
conc_prop_max50NA <- conc_data %>% 
  mutate(Platform = case_when(
    Uniprot %in% ms_proteins_filtered & ! Uniprot %in% olink_proteins_filtered ~ 'MS',
    Uniprot %in% olink_proteins_filtered & ! Uniprot %in% ms_proteins_filtered ~ 'Olink',
    Uniprot %in% ms_proteins_filtered & Uniprot %in% olink_proteins_filtered ~ 'Both',
    !(Uniprot %in% ms_proteins_filtered | Uniprot %in% olink_proteins_filtered) ~ 'Not detected')) %>% 
  mutate(Platform = factor(Platform, levels = c('Both', 'MS', 'Olink', 'Not detected')))


conc_prop_max50NA |>     
  ggplot.histogram(Blood.conc.ngmL, Fill_var = Platform, breaks = 10^(seq(-3, 8, 0.5)),
                   position = position_fill(reverse = T), median_line = F,
                   median_text = F, alpha = 0.9) +
  scale_x_continuous(trans = 'log10', expand = expansion(c(0,0.01)), 
                     breaks = 10^(seq(-3, 8, 1)),
                     labels = log10_format()) +
  scale_y_continuous(expand = expansion(c(0,0.01))) +
  labs(title = '', x = 'Concentration in blood (ng/mL)',
       y = 'Proportion of plasma proteins') +
  scale_color_manual(values = c('Not detected' = TDP_palette[[7]],
                                'Both' = TDP_palette[[13]],
                                TDP_palette[c(1,4)]),
                     aesthetics = c('color', 'fill')) +
  theme_publ()

```

```{r}

ggsave(filename = 'HPA_bloodConc_plasmaProteomeCoverage_max50NA.pdf',
       path = path,
       width = 6.5,
       height = 6.5,
       units = 'cm')

```


## Session info

```{r}

writeLines(capture.output(sessionInfo()), file.path(path, 'session_info.txt'))

```

